#lab_12
Sean Mussenden, Aug 16, 2022

## About this lab

To complete this lab, you need to:
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Overview

There is one main goal of this lab:

* To apply your knowledge of basic web scraping.
* To integrate scraped data with other data pulled from API packages.

You will do this by scraping data from the Maryland State Board of Election website, and connecting it with U.S. Census data pulled from the Tidycensus package to answer questions that can be answered with one or more data sets. 

You can find election results here: https://elections.maryland.gov/elections/2022/general_results/

## Load libraries and settings

**Task**: load the tidyverse, tidycensus, rvest and janitor.  Store your Census api key.

```{r}
#load libraries
library(tidyverse)
library(tidycensus)
library(janitor)
library(rvest)
```

```{r}
#census key
census_api_key("dfcb820a29176622bc44d5c66b77d03b5675937d", install=TRUE, overwrite = TRUE)
readRenviron("~/.Renviron")
```

## Questions

**Q1** Which candidate, Dan Cox or Wes Moore, got the lowest percentage of their votes on Election Day?
What does that suggest?
Write up your answer in a sentence.
**A1**

```{r}
#store govenor results url
gov_race_results <- "https://elections.maryland.gov/elections/2022/general_results/gen_results_2022_1.html"

#read html
gov_race_results <- gov_race_results %>% 
  read_html() %>% 
  html_table()

#clean names and 
gov_race_results <- gov_race_results[[1]] %>% 
  clean_names() %>% 
  mutate(across(c(3,4,5,7,8), parse_number)) %>% 
  mutate(pct_election_day = round(election_day/total*100,2)) %>% 
  slice(-7) %>% 
  arrange(pct_election_day) %>% 
  select(name,pct_election_day)

#Wes Moore got the lowest percentage of their votes from election day, with about 44.3% of his votes coming from election day compared to Dan Cox, who got about 66.3% of his votes from election day. 

#These findings suggests most of Moore's votes came from early voting or mail-in ballots, which aligns with data from other 2022 elections, including the presidental election, that also saw democratic candidates get most of their votes from early or mail-in voting rather than in person on election day. (https://www.pewresearch.org/politics/2023/07/12/voter-turnout-2018-2022/)
  
```

**Q2** How many counties did Chris Van Hollen win in Maryland's Senate race?
**A2**

```{r}
#load senator race data from url
sen_race_results <- "https://elections.maryland.gov/elections/2022/general_results/gen_detail_results_2022_4_1.html"

#read site html
sen_race_results <- sen_race_results %>%
  read_html() %>%
  html_table()

#clean up table and calculate results
sen_race_results <- sen_race_results[[1]] %>%
  clean_names() %>%
  slice(-25) %>%
  clean_names() %>%
  mutate(across(c(2:5), parse_number)) %>%
  rename(chaffee = 2, van_hollen = 3) %>%
  select(jurisdiction, chaffee, van_hollen) %>%
  mutate(
    senate_results = case_when(
      van_hollen > chaffee ~ "Van Hollen",
      chaffee > van_hollen ~ "Chaffee"
    )
  )

sen_race_results %>% 
count(senate_results)

#Chris Van Hollen won 10 counties in Maryland's Senate race
```

**Q3** What percentage of the vote did Chris Van Hollen in Maryland's wealthiest county by median income.
(You must use the Tidycensus package to answer this question).
**A3**

```{r}
#use tidycensus to load variables 
acs_vars <- load_variables("acs5", year=2020)

#find county with the weathiest median income
md <- get_acs(geography = "county",
              variables = c(median_income = "B19013_001"),
              state = "MD",
              year = 2020) %>% 
  arrange(desc(estimate))

#calculate percent of vote
sen_race_results <- sen_race_results %>% 
  mutate(pct_win_hollen = (van_hollen/(chaffee+van_hollen)*100))

sen_race_results <- sen_race_results %>% 
  filter(jurisdiction=="Howard")

#Chris Van Hollen won by about 71% of the vote in Howard County — the wealthiest county by median income in Maryland
```