---
title: Wealth Inequality in Baltimore City and Baltimore County
author: Olivia Borgula
date: 02/29/2024
---

INTRODUCTION

First, load libraries, turn off scientific notation and upload datasets

```{r}
#load libraries
#load libraries
library(tidyverse)
library(janitor)
library(formattable)
library(ggthemes)
library(ggplot2)
library(tidyr)

#turn off scientific notation
options(scipen=999)
```

```{r}
#upload datasets from US Census Bureau 

medians_2012_2022 <- read_csv("/Users/oliviaborgula/Documents/GitHub/olivia_borgula_jour472/data-notebook/data/medians_2012_2022.csv")

poverty_status_2 <- read_csv("/Users/oliviaborgula/Documents/GitHub/olivia_borgula_jour472/data-notebook/data/poverty-3.csv")
```

**Part 1**

What was the median household income in Baltimore City, Baltimore County and Maryland in 2022?

```{r}
#create a pivot table displaying the median household income in each location for 2022
med_summary <- medians_2012_2022 %>%
  filter(year %in% c("2022"))

View(med_summary)

#Findings for 2022 median household income:
#Baltimore City: $55,198
#Baltimore County: $86,526
#Maryland: $94,991
```

**Part 2** 
How many people have poverty status (are below the poverty line) in Baltimore City, Baltimore County and Maryland?

```{r}
#Create a pivot table and filter out values from large census dataset
poverty_number <- poverty_status_2 %>%
  select(`Label (Grouping)`, `Maryland!!Below poverty level!!Estimate`, `Baltimore County, Maryland!!Below poverty level!!Estimate`, `Baltimore city, Maryland!!Below poverty level!!Estimate`)%>%
  filter(`Label (Grouping)` %in% c("Population for whom poverty status is determined"))
View(poverty_number)

#Findings:
#Maryland: 618,372 people
#Baltimore County: 81,052 people
#Baltimore City: 127,135 people

```

**Part 3** 
What percent of people have poverty status compared to the total population in Maryland, Baltimore City and Baltimore County?

```{r}

reshaped_poverty <- poverty_pivot %>%
  select




#Findings:
#Maryland: 10.3% of population is poverty status
#Baltimore County: 9.8% of population is poverty status
#Baltimore City: 23.0%% of population is poverty status 

```

**Part 4** 
Visualize this data with bar chart

```{r}
#Make a bar chart with new column
reshaped_data %>%
  ggplot(aes(x=Location, y= Percent, fill = "blue")) +
  theme_minimal() +
  geom_col(position = "dodge")+
  theme(legend.position = "none") +
  geom_text(aes(label=Percent), vjust = -.5, size = 2.5) +
    labs(
    title="Percent of People Below Poverty Line in Maryland, Baltimore City and Baltimore County",
    x = "Location",
    y = "Percent",
    caption = "Source: U.S. Census Bureau ")
```

Determine the demographic makeup of people living below the poverty line in 2022 in Baltimore City vs. Baltimore County

```{r}

```

Find the percent change in median household income from 2012 to 2022 in Baltimore County vs. Baltimore City

```{r}
#MAKE TABLE
#First make a pivot table to more easily identify values
decade_values <- medians_2012_2022 %>% 
  filter(year %in% c("2022", "2012")) 

#Find percent change for Baltimore City between 2012 and 2022
percent((55198 - 39241)/39241)

#Find percent change for Baltimore County between 2012 and 2022
percent((86526 - 62444)/62444)

```

Visualize the median household income in Baltimore City and Baltimore County over that same time period

```{r}
ggplot(medians_2012_2022, aes(x=year)) +
  geom_line(aes(y=balt_city), color = "blue", show.legend = TRUE) +
  geom_line(aes(y=balt_county), color = "green")+
  geom_line(aes(y=maryland), color = "red")+
  labs(title = "Median Household Income from 2012-2022", 
       caption = "Source: U.S. Census Bureau",
       y="Income",
       x="Year")
```
