---
title: Wealth Inequality in Baltimore City and Baltimore County
author: Olivia Borgula
date: 02/29/2024
---

INTRODUCTION

First, load libraries, turn off scientific notation and upload datasets
```{r}
#load libraries
library(tidyverse)
library(janitor)
library(formattable)
library(ggthemes)

#turn off scientific notation
options(scipen=999)

#upload datasets 
baltcity <- read_csv("/Users/oliviaborgula/Documents/GitHub/olivia_borgula_jour472/data-notebook/baltcity_income_clean.csv")
poverty_status <- read_csv("/Users/oliviaborgula/Documents/GitHub/olivia_borgula_jour472/data-notebook/2022-census/poverty-status.csv")
data7 <- read_csv("/Users/oliviaborgula/Documents/GitHub/olivia_borgula_jour472/data-notebook/2022-census/data-7.csv")

View(data7)
```

Identify median household income data in Baltimore City and in to Baltimore County in 2023 
```{r}
summary()
```

Number of individuals, families and children living below the poverty line in 2023 in each county
```{r}

```

Determine the demographic makeup of Baltimore city families living below the poverty line
```{r}

```

Compare each state with the median household income for the entire state
```{r}

```

Determine percent change in median household income between 2010 and 2020 for Baltimore City and Baltimore County 
```{r}

```

Create a line graph with median household income in Baltimore City and Baltimore County each year from 2010 to 2020
```{r}

```

