---
title: "Olivia Borgula Data Analysis Project"
date: 4/25/24
format: html
editor: visual
---

**INTRODUCTION**

**SET UP**
```{r}
#libraries and settings
library(tidyverse)
library(janitor)
library(formattable)
library(dplyr)
library(rvest)
library(tidyverse)
library(tidycensus)

#turn off scientific notation
options(scipen=999)
```

```{r}
#upload census key
census_api_key("dfcb820a29176622bc44d5c66b77d03b5675937d", install=TRUE, overwrite = TRUE)
readRenviron("~/.Renviron")
```

**Q1: In the top 10 areas with the highest incarceration rates, how many people registered to vote in the 2020 election? What about in the 10 areas with the lowest incarceration rate?**

Part 1: in the top 10 areas with the highest incarceration rates, how many people registered to vote in the 2020 election? 
```{r}
#webscrape data table from prison policy initiative website 
balt_incarceration <- "https://www.prisonpolicy.org/origin/md/2020/baltimore_csa.html"
balt_incarceration <- balt_incarceration %>% 
  read_html() %>% 
  html_table()

balt_incarceration <- balt_incarceration[[1]]%>%
  clean_names()%>%
  rename(neighborhood = 1, num_prison = 2, x2020_census_pop = 3, x2020_total_pop = 4, rate_per_x100000 = 5)

balt_incarceration <- balt_incarceration %>%
  mutate(rate_per_x100000 = as.numeric(gsub(",", "", rate_per_x100000)))%>%
    arrange(desc(rate_per_x100000))
#upload census data for voter registration in the 2020 election


#join balt_incarceration table and census data table


#sort by top 10 incarcerated areas and find how many people registered to vote 


#sort by top 10 lowest incarcerated areas and find how mnay people registered to vote
```


**Q2: In each of the top 10 areas with the lowest voter turnout, what’s the percentage of people with a member of their family incarcerated? How does that compare to areas  of high voter turnout?**
```{r}
#upload data for people with members of their family incarcerated 
```


**Q3: How many people registered to vote in the 2020 election in the top five areas in Baltimore City with the highest median household income? What about in the five lowest areas? **
```{r}
#upload census data for median household income in baltimore in 2020


#join median household income table and voter registration data for 2020


#find top five areas with highest median household income 


#find top five areas with lowest median household income 
```


**Q4: What race in Baltimore City had the largest disparity in number of registered voters and number of reported voting in the 2020 election? Which had the lowest?**
```{r}
v17 <- load_variables(2017, "acs5", cache = TRUE)

View(v17)
```

