---
title: "sillyness"
format: html
editor: visual
---

```{r, results='asis'}
cat('<div class="flourish-embed flourish-map" data-src="visualisation/17809409"><script src="https://public.flourish.studio/resources/embed.js"></script></div>')
```
1. Which five Baltimore City community statistical areas had the lowest voter turnout percentage in the 2020 general election? Which five had the highest?

2. Are there correlations between the Baltimore City 2020 general election voter turnout and factors including incarceration rate, business activity, wealth and internet access?

3. In 2020, how many registered voters were democrats and how many were republicans? How does that compare to the rest of Maryland? 
- sum of dem column, sum of republican column in precinct_voters
- visualize
- which csa has the highest concentration of dems/republicans? - spatial join -> add up total in each csa -> divide by population of csa
- how does concentration of dems/republicans compare to race? 

4. Which race had the least amount of registered voters as a proportion of their population in the 2020 general election?
- ?

5. How has voter registration in Baltimore City changed from 2020 to 2024? 
- clean data -> percent of the city registered 

```{r}
#voting data by voting precinct in baltimore city in August 2020 
x2020_precinct_voters <- read_csv("data/cleaned_precinct_x2020.csv")
#remove blank rows
x2020_precinct_voters <- x2020_precinct_voters[complete.cases(x2020_precinct_voters), ]

x2020_precinct_voters <- x2020_precinct_voters[-111, ]%>%
  rename(total_reg = total)
```


**In 2020, which Baltimore City voting precinct had the highest concentration of democrats? Which had the highest concentration of republicans? Is there any correlation between race and party affiliation?**

For this question, I'm looking at voting precincts rather than community statistical areas. Voting precincts is a subdivision of an electoral district. Typically, each contains one polling place where everyone within that precinct goes to vote. Community statistical areas are clusters of recognizable neighborhoods that are more useful for comparing things such as wealth, internet access and incarceration, but for voting data, the most accurate is at the precinct-level. 

Let's start with finding concentrations of democrats/republicans in voting precincts
```{r}
#create new column for percent of whole for dem/rep
x2020_precinct_voters <- x2020_precinct_voters%>%
  mutate(pct_dem = (dem / total_reg)*100)%>%
  mutate(pct_rep = (rep / total_reg)*100)
```
This won't make much sense without a visualization:
```{r}
cat('')
```

x2020-x2024_precinct_voters
```{r}
#voting data by voting precinct in baltimore city in 2024 
x2024_precinct_voters <- read_csv("data/tabula-2024-precinct-voter-counts.csv")
#remove blank rows
x2024_precinct_voters <- x2024_precinct_voters[complete.cases(x2024_precinct_voters), ]

#voting data by voting precinct in baltimore city in 2023 
x2023_precinct_voters <- read_csv("data/tabula-2023_reg.csv")

#voting data by voting precinct in baltimore city in 2022 
x2022_precinct_voters <- read_csv("data/tabula-2022_dec.csv")

#voting data by voting precinct in baltimore city in 2021 
x2021_precinct_voters <- read_csv("data/tabula-2021_reg.csv")

#voting data by voting precinct in baltimore city in August 2020 
x2020_precinct_voters <- read_csv("data/cleaned_precinct_x2020.csv")
#remove blank rows
x2020_precinct_voters <- x2020_precinct_voters[complete.cases(x2020_precinct_voters), ]

x2020_precinct_voters <- x2020_precinct_voters[-111, ]%>%
  rename(total_reg = total)%>%
  clean_names()
```

6. total_22 - data with the number of democratic, republican and total voters from the 2022 gubernatorial race by voting precinct. I used Tabula to extract the data from this website and then upload it as a CSV. https://boe.baltimorecity.gov/sites/default/files/2022-12-28%201306%20-%20PUBLISHED%20-%20On%20WELCOME,%20MAIN%20MENU,%20VOTER%20INFORMATION,%20VOTER%20REGISTRATION,%20REGISTRATION%20FIGURES%20Page%20-%202022-12-28%20-%20City%20Wide%20Precinct%20&%20Wards.pdf

7. total_2018 - data with the number of democratic, republican and total voters from the 2018 gubernatorial race by voting precinct. I used Tabula to extract the data from this website and then upload it as a CSV. 
https://boe.baltimorecity.gov/sites/default/files/congressional%20voter%20totals%20-%202018-2-27.pdf

**Q**
Find the number of democratic voters and republican voters in each voting precinct in 2022 and 2018 as a share of the total voters in each precinct. 

2022
```{r}
total_22 <- total_22%>%
  mutate(pct_dem = (dem / total)*100)%>%
  mutate(pct_rep = (rep / total)*100)
```

2018
```{r}
total_2018 <- total_2018 %>%
  mutate(dem = as.numeric(dem),  
         rep = as.numeric(rep),  
         total = as.numeric(total))   

total_2018 <- total_2018 %>%
  mutate(pct_dem = (dem / total) * 100,   
         pct_rep = (rep / total) * 100)   

```


Voter precinct data doesn't make much sense without a visualization, but is more detailed than Baltimore City community statistical area. So let's visualize it.
```{r}
cat('')

dem_gov_flour<- dem_gov %>%
  select(election_district_precinct, sum)
```


total_22
```{r}
#upload data
total_22 <- read_csv("data/total_22_clean.csv")%>%
  clean_names

#clean table
total_22 <- total_22%>%
  rename(precinct = 1)%>%
  select(precinct, total, dem, rep)%>%
  na.omit
```

total_2018
```{r}
#upload data
total_2018 <- read_csv("data/total_2018_clean.csv")%>%
  clean_names()

#clean table
total_2018 <- total_2018%>%
  rename(precinct = 1)%>%
  select(precinct, total, dem, rep)%>%
  na.omit
```
