---
title: "Olivia Borgula Data Analysis Project"
date: 4/25/24
format: html
editor: visual
---
**NOTE**
I'm submitting this milestone at 11:30 pm on 4/25, but plan to make the following additions tomorrow evening:
- finishing cleaning and uploading data for 2024 voting numbers by voting precinct in Baltimore City 
- uploading a dataset I found after cleaning that has one column for neighborhood and one column for voting precinct numbers, which I will then join with other data tables for further analysis 


**INTRODUCTION**
This data analysis project will examine how incarceration, wealth and race intersect and play a role in voter registration in Baltimore City — an especially newsworthy topic ahead of elections season this November. Specifically, I will ask these five questions to guide my findings:

1. In the top 10 areas in Baltimore City with the highest incarceration rates, how many people registered to vote in August 2020? What about in the 10 areas with the lowest incarceration rate?

2. In each of the top 10 areas with the lowest voter turnout in 2020, how many people were incarcerated? How does that compare to areas of high voter turnout?

3. How many people registered to vote in August 2020 in the top five neighborhoods in Baltimore City with the highest median household income? What about in the five lowest areas?

4. What race in Baltimore City had the largest disparity in number of registered voters and number of reported voting in the 2020 election? Which had the lowest?

5. How have voter registration rates for people in highly-incarcerated areas changed in the past four years? 

**THE DATA**
The core data sets include: 
1. x2020_precinct_voters — voter registration counts in August 2020 by voting precinct in Baltimore City via the city's voting information page

2. x2024_precinct_voters - voter registration counts in January 2024 by voting precinct in Baltimore City via the city's voting information page 
https://boe.baltimorecity.gov/registration-figures <- for above 2 datasets

3. demo_voting_x2020 — voter demographic information by voting precinct in 2020 in Baltimore City via the Maryland Department of Planning
https://planning.maryland.gov/Redistricting/Documents/2020data/GreenReport.pdf

4. incarceration_wealth — a data set formed by a join of two datasets. One is baltcity_income, given to us in past labs and containing information about the median household income, and balt_incarceration, a data table scraped from a Prison Policy Initiative website with information about the incarceration rate in different neighborhoods in 2020. 
prison policy initiative website: https://www.prisonpolicy.org/origin/md/2020/baltimore_csa.html

**SET UP**

```{r}
#libraries and settings
library(tidyverse)
library(janitor)
library(formattable)
library(dplyr)
library(rvest)
library(tidyverse)
library(tidycensus)

#turn off scientific notation
options(scipen=999)
```

```{r}
#upload census key
census_api_key("dfcb820a29176622bc44d5c66b77d03b5675937d", install=TRUE, overwrite = TRUE)
readRenviron("~/.Renviron")
```

```{r}
#upload cvs and start cleaning data 
#voting data by voting precinct in baltimore city in 2024 
x2024_precinct_voters <- read_csv("data/tabula-2024-precinct-voter-counts.csv")
#remove blank rows
x2024_precinct_voters <- x2024_precinct_voters[complete.cases(x2024_precinct_voters), ]

#voting data by voting precinct in baltimore city in August 2020 
x2020_precinct_voters <- read_csv("data/cleaned_precinct_x2020.csv")
#remove blank rows
x2020_precinct_voters <- x2020_precinct_voters[complete.cases(x2020_precinct_voters), ]
  

#income in baltimore city
baltcity_income <- read_csv("data/baltcity_income.csv")%>%
  clean_names()

#voting presidential voter registration counts by counties in maryland
x2020_county_voting <- read_csv("data/county_voter_x2020_clean.csv")
x2020_county_voting <- x2020_county_voting[, -c(2, 4, 13)]

```

**Q1: In the top 10 areas with the highest incarceration rates, how many people registered to vote in the 2020 election? What about in the 10 areas with the lowest incarceration rate?**

Part 1: in the top 10 areas with the highest incarceration rates, how many people registered to vote in the 2020 election?

```{r}
#webscrape data table from prison policy initiative website 
balt_incarceration <- "https://www.prisonpolicy.org/origin/md/2020/baltimore_csa.html"
balt_incarceration <- balt_incarceration %>% 
  read_html() %>% 
  html_table()

balt_incarceration <- balt_incarceration[[1]]%>%
  clean_names()%>%
  rename(neighborhood = 1, num_prison = 2, x2020_census_pop = 3, x2020_total_pop = 4, rate_per_x100000 = 5)

balt_incarceration <- balt_incarceration %>%
  mutate(rate_per_x100000 = as.numeric(gsub(",", "", rate_per_x100000)))%>%
    arrange(desc(rate_per_x100000))

#top ten table
top_balt_incarcerated <- balt_incarceration %>%
  select(neighborhood, rate_per_x100000)%>%
  slice_max(rate_per_x100000, n = 10)
```

Part 2: what about the top ten areas with the lowest?
```{r}
#bottom ten table of incarceration rates
bottom_balt_incarcerated <- balt_incarceration %>%
  select(neighborhood, rate_per_x100000)%>%
  slice_min(rate_per_x100000, n = 10)
```

**Q2: In each of the top 10 areas with the lowest voter turnout, how many people were incarcerated? How does that compare to areas of high voter turnout?**

```{r}

```


**Q3: How many people registered to vote in the 2020 election in the top five areas in Baltimore City with the highest median household income? What about in the five lowest areas?**

```{r}
#join baltcity_income and balt_incarceration
incarceration_wealth <- baltcity_income %>%
  inner_join(balt_incarceration, by="neighborhood")

incarceration_wealth

```


**Q4: What race in Baltimore City had the largest disparity in number of registered voters and number of reported voting in the 2020 election? Which had the lowest? **

```{r}

```

**Q5: How have voter registration rates for people in highly-incarcerated areas changed in the past four years?**
For this I plan to do data visualization 

