---
title: "Data Analysis Project Milestone 4"
name: "Olivia Borgula"
date: 05/04/2024
format: html
editor: visual
---

**INTRODUCTION** 
This data analysis project will examine how different factors, such as incarceration, wealth and race, play a role in Baltimore City's 2020 general election voter turnout. It will also examine how voter registration has changed from 2020 to 2024, an especially newsworthy topic ahead of election season this November. Specifically, I will ask these five questions to guide my findings:

1. Which five Baltimore City community statistical areas had the lowest voter turnout percentage in the 2020 general election? Which five had the highest?

2. Are there correlations between the Baltimore City 2020 general election voter turnout and factors including incarceration rate, business activity, wealth and internet access?

3. How has voter turnout in Baltimore City changed from 2010 to 2020? Which CSA had the biggest difference between the 2012 and 2020 general election?

**THE DATA** 
Note: if you click on the Baltimore Neighborhood Indicators Alliance links, make sure the year is set to 2020 (Compare Years -> 2020 -> Update Your Selections) and the corresponding indicator on the left-hand side is clicked. It doesn't automatically apply these when I copy the link

1. pct_vote_csa_clean - data from the Baltimore City Board of Elections compiled by the Baltimore Neighborhood Indicators Alliance that has the percent of people over the age of 18 who voted in the 2020 general election in each Baltimore City community statistical area. https://bniajfi.org/indicators/Sustainability/voted/2020. I exported the data from the website as an excel sheet and then saved it as a csv. 

2. incarceration_x2020 - a data table scraped from a Prison Policy Initiative website with information about the incarceration rate in each Baltimore City community statistical areas in 2020. Rate per 100,000 is the number of imprisoned people divided by the total CSA population and then multiplied by 100,000. It allows ready comparison of the frequency of imprisonment between each CSA of different population sizes https://www.prisonpolicy.org/origin/md/2020/baltimore_csa.html

3. biz - data from InfoUsa compiled by the Neighborhood Indicators Alliance that has the number of businesses per 1,000 residents by community statistical area. https://bniajfi.org/indicator/Workforce%20And%20Economic%20Development/?chkYears[]=2020. I exported the data from the website as an excel sheet and then saved it in excel as a csv. 

4. baltcity_income - data from the US Census Bureau compiled from the Neighborhood Indicators Alliance that has the median household income in each Baltimore City community statistical area. https://bniajfi.org/indicator/census%20demographics/. I exported the data from the website as an excel sheet and then saved it as a csv. 

5. internet - data from US Census Bureau compiled by the Neighborhood Indicators Alliance that has the percent of households with no internet at home. https://bniajfi.org/indicators/Workforce%20and%20Economic%20Development/nohhint/2020. I exported the data from the website as an excel sheet and then saved it in excel as a csv.

6. voting_2012 through voting_2022 - data from the Baltimore City Board of Elections compiled by the Neighborhood Indicators Alliance that has the voter turnout rate for each Baltimore City statistical area

**SET UP**
```{r}
#libraries and settings
library(tidyverse)
library(janitor)
library(formattable)
library(dplyr)
library(rvest)
library(sf)

#turn off scientific notation
options(scipen=999)
```

**UPLOAD DATA**
pct_vote_csa
This dataset has a row for all of Baltimore City, which is not one of the official 55 CSAs, so I will remove that row
```{r}
#upload csv and clean names/rows
pct_vote_csa <- read_csv("data/pct_vote_csa_clean.csv")

pct_vote_csa <- pct_vote_csa%>%
  clean_names()%>%
  rename(csa = community, pct_voters = x2020_data)%>%
  filter(csa != "Baltimore City")
``` 

incarceration_x2020
```{r}
#webscrape data table from prison policy initiative website 
incarceration_x2020 <- "https://www.prisonpolicy.org/origin/md/2020/baltimore_csa.html"
incarceration_x2020 <- incarceration_x2020 %>% 
  read_html() %>% 
  html_table()
incarceration_x2020 <- incarceration_x2020[[1]]%>%
  clean_names()%>%
  rename(csa = 1, rate_per_x100000 = 5)%>%
  select(csa, rate_per_x100000)
```

baltcity_income
This dataset also has a row for all of Baltimore City, which is not one of the official 55 CSAs, so I will remove that row
```{r}
baltcity_income <- read_csv("data/baltcity_income.csv")%>%
  clean_names()%>%
  rename(csa = community, median_income = x2020_data)%>%
  filter(csa != "Baltimore City")
```

biz
This dataset also has a row for all of Baltimore City, which is not one of the official 55 CSAs, so I will remove that row
```{r}
biz <- read_csv("data/smallbiz_csa.csv")

biz <- biz%>%
  clean_names()%>%
  rename(csa = community, num_biz_x1000 = x2020_data)%>%
  filter(csa != "Baltimore City")
```

internet
This dataset also has a row for all of Baltimore City, which is not one of the official 55 CSAs, so I will remove that row
```{r}
internet <- read_csv("data/internet.csv")

internet <- internet%>%
  clean_names()%>%
  rename(csa = community, pct_no_internet = x2020_data)%>%
  filter(csa != "Baltimore City")
```

voting_2012 through voting_2020
These datasets also has a row for all of Baltimore City, which is not one of the official 55 CSAs, so I will remove that row
```{r}
#voter turnout in the 2010 election
voting_2010 <- read_csv("data/voting_2010.csv")%>%
  clean_names()%>%
  rename(csa = community, pct_voters = x2010_data)%>%
  filter(csa != "Baltimore City")

#voter turnout in the 2012 election
voting_2012 <- read_csv("data/voting_2012_clean.csv")%>%
  clean_names()%>%
  rename(csa = community, pct_voters = x2012_data)%>%
  filter(csa != "Baltimore City")

#voter turnout in the 2014 election
voting_2014 <- read_csv("data/voting_2014.csv")%>%
  clean_names()%>%
  rename(csa = community, pct_voters = x2014_data)%>%
  filter(csa != "Baltimore City")

#voter turnout in the 2016 election
voting_2016 <- read_csv("data/voting_2016.csv")%>%
  clean_names()%>%
  rename(csa = community, pct_voters = x2016_data)%>%
  filter(csa != "Baltimore City")

#voter turnout in the 2018 election
voting_2018 <- read_csv("data/voting_2018.csv")%>%
  clean_names()%>%
  rename(csa = community, pct_voters = x2018_data)%>%
  filter(csa != "Baltimore City")

#voter turnout in the 2022 ele
voting_2016 <- read_csv("data/voting_2016.csv")%>%
  clean_names()%>%
  rename(csa = community, pct_voters = x2016_data)%>%
  filter(csa != "Baltimore City")
```


**Q1: Which five Baltimore City community statistical areas had the lowest voter turnout percentage in the 2020 general election? Which five had the highest?** 

```{r}
#use slicemax/min to make new tables with top and bottom five csas by voter turnout
pct_vote_csa%>%
  slice_min(pct_voters, n = 5)
#Southwest Baltimore: 41.1%, Madison/East End: 41.6%, Brooklyn/Curtis Bay/Hawkins Point: 42.7%, Sandtown-Winchester/Harlem Park: 44.1% and Poppleton/The Terraces/Hollins Market: 46.1%

pct_vote_csa%>%
  slice_max(pct_voters, n = 5)
#Greater Roland Park/Poplar Hill: 80.1%, Mount Washington/Coldspring: 77%, South Baltimore: 75%, North Baltimore/Guilford/Homeland: 74.6% and Medfield/Hampden/Woodberry/Remington: 72.9%
```

Let's visualize this! 
I created a custom map in flourish by uploading a geojson file with the Baltimore City CSA boundaries, which I found through Open Baltimore. https://data.baltimorecity.gov/datasets/9c96ae20e6cc41258015c2fd288716c4/explore
I then uploaded my csv for pct_vote_csa to fill in the data points. 

```{r, results='asis'}
cat('<div class="flourish-embed flourish-map" data-src="visualisation/17796037"><script src="https://public.flourish.studio/resources/embed.js"></script></div>')
```

Looking at this visualization, it's clear the areas with the highest voter turnout are in the middle areas of Baltimore, especially in the Inner Harbor area stretching north to Homeland and Guilford, with east and west Baltimore showing significantly less voter turnout. Sixteen CSAs had a voter turnout of less than 50%, and they were all concentrated in similar areas. 

**Q2: Are there correlations between the Baltimore City 2020 general election voter turnout and factors including incarceration rate, business activity, wealth and internet access?**

Let's compare all these factors in the five CSAs with the highest and lowest voter turnout rates

I'll start with incarceration rate and voter turnout
```{r}
#join incarceration_x2020 and pct_vote_csa datasets
incarceration_voting <- inner_join(incarceration_x2020, pct_vote_csa, by = "csa")

incarceration_voting %>%
  slice_min(pct_voters, n = 5)

incarceration_voting %>%
  slice_max(pct_voters, n = 5)

#Incarceration rate per 100,000 people in five areas of lowest voter turnout:
#Southwest Baltimore: 2,223, Madison/East End: 2,528, Brooklyn/Curtis Bay/Hawkins Point: 1,204, Sandtown-Winchester/Harlem Park: 2,563 and Poppleton/The Terraces/Hollins Market: 1,623

#Incarceration rate per 100,000 people in five areas of highest voter turnout:
#Greater Roland Park/Poplar Hill: 14, Mount Washington/Coldspring: 102, South Baltimore: 97, North Baltimore/Guilford/Homeland: 93 and Medfield/Hampden/Woodberry/Remington: 225
```

Let's visualize this!
```{r, results='asis'}
cat('<div class="flourish-embed flourish-map" data-src="visualisation/17809409"><script src="https://public.flourish.studio/resources/embed.js"></script></div>')
```
This visualization shows that there is a correlation between incarceration rate and voter turnout. Looking at the map of voter turnout next to the map of incarceration rates, it's clear that the two are generally inverses of on another, with areas in the north and middle of Baltimore having a low incarceration rate and high voter turnout and east and west Baltimore having a higher incarceration rate and low voter turnout.  

Another visualization. This time, a scatterplot
```{r, results='asis'}
cat('<div class="flourish-embed flourish-scatter" data-src="visualisation/17795149"><script src="https://public.flourish.studio/resources/embed.js"></script></div>')
```
This scatterplot shows the negative relationship between incarceration rate and voter turnout. As the rate of incarceration per 100,000 people increases, the percent of registered voters decreases. 

Wealth and voter turnout rate
```{r}
#join baltcity_income and pct_vote_csa 
wealth_voting <- inner_join(pct_vote_csa, baltcity_income, by = "csa")

wealth_voting %>%
  slice_min(pct_voters, n = 5)

wealth_voting %>%
  slice_max(pct_voters, n = 5)

#median household income in areas of lowest voter turnout (rounded to nearest dollar):
#Southwest Baltimore: $29,768, Madison/East End: $37,303, Brooklyn/Curtis Bay/Hawkins Point: $32,598, Sandtown-Winchester/Harlem Park: $26,690 and Poppleton/The Terraces/Hollins Market: $23,374

#median household income in areas of highest voter turnout:
#Greater Roland Park/Poplar Hill: $125,869, Mount Washington/Coldspring: 86,715, South Baltimore: $124,827, North Baltimore/Guilford/Homeland: $114,247 and Medfield/Hampden/Woodberry/Remington: $72,757
```

Let's visualize this one, too
```{r, results='asis'}
cat('<div class="flourish-embed flourish-map" data-src="visualisation/17824036"><script src="https://public.flourish.studio/resources/embed.js"></script></div>')
```
Looking at this map shows that areas of higher median household income are concentrated in the middle of Baltimore City and in the northern areas. This is a similiar pattern to the voter turnout rate map.

Another visualization
```{r, results='asis'}
cat('<div class="flourish-embed flourish-scatter" data-src="visualisation/17824265"><script src="https://public.flourish.studio/resources/embed.js"></script></div>')
```
This scatterplot shows the positive relationship between median household income and voter turnout. As the median household income increases, the percent of registered voters decreases.

Number of businesses per 1,000 people and voter turnout  
```{r}
#join biz and pct_vote_csa dataset
biz_voting <- inner_join(biz, pct_vote_csa, by = "csa")

biz_voting %>%
  slice_min(pct_voters, n = 5)

biz_voting %>%
  slice_max(pct_voters, n = 5)

#number of businesses per 1,000 people in five areas of lowest voter turnout:
#Southwest Baltimore: 18.4, Madison/East End: 19.2, Brooklyn/Curtis Bay/Hawkins Point: 13.4, Sandtown-Winchester/Harlem Park: 10.4 and Poppleton/The Terraces/Hollins Market: 14.2

#number of businesses per 1,000 people in five areas of highest voter turnout:
#Greater Roland Park/Poplar Hill: 33.5, Mount Washington/Coldspring: 31.4, South Baltimore: 21.9, North Baltimore/Guilford/Homeland: 16.3 and Medfield/Hampden/Woodberry/Remington: 35.4
```
There doesn't seem to be a significant correlation between number of businesses per 1,000 residents and voter turnout. While looking at just the top 5 top and bottom areas of high/low voter turnout seems like areas of lower voter turnout have less businesses, there are quite a few outliers, including Oldtown/Middle East, which was one of the lowest voter turnout areas, but had the second highest amount of businesses per 1,000 residents. This could be for a variety of reasons. For one, many of the areas that have high voter turnout are in wealthier, suburban areas with less concentration of businesses. 

Percent of households without internet and voter turnout 
```{r}
#join internet and pct_vote_csa dataset
internet_voting <- inner_join(internet, pct_vote_csa, by = "csa")

internet_voting %>%
  slice_min(pct_voters, n = 5)

internet_voting %>%
  slice_max(pct_voters, n = 5)

#Percentage of homes without internet access in the five areas of lowest voter turnout:
#Southwest Baltimore: 27.6%, Madison/East End: 16.7%, Brooklyn/Curtis Bay/Hawkins Point: 30.4%, Sandtown-Winchester/Harlem Park: 38.6% and Poppleton/The Terraces/Hollins Market: 25.1%

#Percentage of homes without internet access in the five areas of highest voter turnout:
#Greater Roland Park/Poplar Hill: 3.4%, Mount Washington/Coldspring: 8.9%, South Baltimore: 3.5%, North Baltimore/Guilford/Homeland: 6.1% and Medfield/Hampden/Woodberry/Remington: 9.1%
```

Let's do another visualization
```{r, results='asis'}
cat('<div class="flourish-embed flourish-scatter" data-src="visualisation/17824515"><script src="https://public.flourish.studio/resources/embed.js"></script></div>')
```
This scatterplot shows the negative relationship between percent of households without internet access and voter turnout. As the percent of households without internet increases, the percent of registered voters decreases.

Just to be really sure, let's compute some correlation coefficients. Values close to 1 or -1 indicate a strong positive or negative correlation, while values closer to zero indicate a weak correlation. 

This is not something I would put in an actual article since it's a more complex concept for readers to digest, but it's useful for checking my work and determining whether these factors are worth investigating at all.

Correlation between incarceration rate and voter turnout.
```{r}
#convert all values to numeric for calculation
incarceration_voting$rate_per_x100000 <- gsub("[, ]", "", incarceration_voting$rate_per_x100000)
incarceration_voting$rate_per_x100000 <- as.numeric(incarceration_voting$rate_per_x100000)

#compute correlation coefficient
cc_incarceration <- cor(incarceration_voting$pct_voters, incarceration_voting$rate_per_x100000, method = "pearson", use = "complete.obs")

cc_incarceration

#a correlation coefficient of -0.82 indicates a strong negative correlation, meaning areas of lower incarceration rates likely have higher voter turnout rates. 
```

Correlation between number of businesses per 1000 residents and voter turnout 
```{r}
#compute correlation coefficient
cc_biz <- cor(biz_voting$pct_voters, biz_voting$num_biz_x1000, method = "pearson", use = "complete.obs")

cc_biz

#There is little to no correlation between the amount of businesses and voter turnout, as indicated by a correlation coefficient of -0.02
```

Correlation between median household income and voter turnout 
```{r}
cc_wealth <- cor(wealth_voting$pct_voters, wealth_voting$median_income, method = "pearson", use = "complete.obs")

cc_wealth

#A correlation coefficient of 0.81 indicates a strong positive correlation between voter turnout and median household income
```

Correlation between percentage of households without internet access and voter turnout
```{r}
cc_internet <- cor(internet_voting$pct_voters, internet_voting$pct_no_internet, method = "pearson", use = "complete.obs")

cc_internet

#A correlation coefficient of -0.71 indicates a strong negative correlation between voter turnout and median household income. It's not as strong as the other two, but is still considered strong 
```

**Q3: How has voter turnout in Baltimore City changed from 2010 to 2020? Which CSA had the biggest difference between the 2012 and 2020 general election?**

Let's look at community statistical areas first 
```{r}
#rename row from "&" to "and" to match up with the rest of the voting data used later
pct_vote_csa$csa <- gsub("&", "and", pct_vote_csa$csa)

#make one dataset for combined 2012 and 2020
voting_x2012_x2020 <- inner_join(voting_2012, pct_vote_csa, by = "csa")

voting_x2012_x2020 <- voting_x2012_x2020 %>%
  rename(x2012_turnout = 2, x2020_turnout = 3) %>%
  mutate(pct_change = ((x2020_turnout - x2012_turnout) / x2012_turnout) * 100)

#Orangeville/East Highlandtown had the biggest change, with an almost 65% increase in voter turnout 
```

I'm going to make one big dataset to visualize this 
```{r}
voting_all <- inner_join(voting_2010, voting_2012, by = "csa") %>%
                 inner_join(voting_2014, by = "csa") %>%
                 inner_join(voting_2016, by = "csa") %>%
                 inner_join(voting_2018, by = "csa") %>%
                 inner_join(pct_vote_csa, by = "csa")

voting_all <- voting_all%>%
  rename(x)
```

