---
title: "Data Analysis Project"
name: "Olivia Borgula"
date: "05/04/2024"
format: html
editor: visual
---

This data analysis project examines how different factors — including incarceration, wealth, internet access, business activity and race — impacted Baltimore City's 2020 general election voter turnout.

It explores how voter turnout has changed in the last 10 years across midterm and general elections and compares political party voting patterns in the city.

**Findings**

-   Baltimore City residents in areas of high incarceration were less likely to vote in the 2020 general election. As the rate of incarceration per 100,000 people increased, voter turnout percentage decreased.

-   Areas of low internet access saw a similar pattern. In Southwest Baltimore, which had the lowest voter turnout in the 2020 general election, almost 28% of households did not have internet access. But in the Greater Roland Park and Poplar Hill area, which had the highest voter turnout percentage, only about 3% of households did not have internet.

-   Median household income, on the other hand, had a positive relationship to voter turnout. Wealthier areas had a significantly higher voter turnout percentage. Of the 16 areas with less than 50% of their voting-age population casting their ballots, only one had a median household income of more than \$42,000.

-   Overall voter turnout in the city has significantly decreased since 2008, dropping from 68% to 58% in that time. 2020 saw the first rise in the decade-long period, with most election turnout numbers remaining below the Maryland average. In 2020, areas that had the most democratic-registered voters were also the areas with the highest concentration of Black residents and the lowest voter turnouts.

Data comes from the following places:

-   <a href="https://bniajfi.org/indicator/census%20demographics/" target="_blank" style="color: black;"> Prison Policy Initiative table</a> web scraped from the organization's website with information about the incarceration rate in each Baltimore City community statistical areas in 2020. Its figure indicating the rate per 100,000 people allows comparison imprisonment levels between each CSA of different population sizes.

-   I used Tabula to get a table from <a href = "https://boe.baltimorecity.gov/registration-figures" target="_blank" style="color: black;"> this PDF </a> from the Baltimore City Board of Elections to show voter registration numbers by political party in each voting precinct in Maryland.

-   I used Tabula to get a table from <a href = "https://planning.maryland.gov/Redistricting/Documents/2020data/GreenReport.pdf" target="_blank" style="color: black;">this PDF</a> from the Maryland Department of Planning with a demographic breakdown of Maryland voting precinct populations in the 2020 general election.

-   <a href="https://bniajfi.org/indicators/Sustainability/voted/2020" target="_blank" style="color: black;">Baltimore City Board of Elections data</a> compiled by the Baltimore Neighborhood Indicators Alliance. It indicates the percent of people over the age of 18 who voted in the 2020 general election in each Baltimore City community statistical area.

-   <a href="https://bniajfi.org/indicator/Workforce%20And%20Economic%20Development/?chkYears[]=2020" target="_blank" style="color: black;"> InfoUSa data </a> compiled by the Neighborhood Indicators Alliance that has the number of businesses per 1,000 residents in each Baltimore City community statistical area.

-   <a href="https://bniajfi.org/indicator/census%20demographics/" target="_blank" style="color: black;"> US Census Bureau data </a> with the median household income in each Baltimore City community statistical area.

-   <a href="https://bniajfi.org/indicators/Workforce%20and%20Economic%20Development/nohhint/2020" target="_blank" style="color: black;"> US Census Bureau data </a> with the percent of households with no internet at home by community statistical area.

-   <a href= "https://bniajfi.org/indicators/Sustainability/voted/2020" target="_blank" style="color: black;"> Baltimore City Board of Elections data </a> that has the voter turnout rate for each Baltimore City community statistical area.

-   To compare voting turnout over time, I web scraped data from <a href= "https://elections.maryland.gov/elections/presidential_elections/turnout/general_comparison.html" target="_blank" style="color: black;"> this site </a>, which has the Baltimore City voter turnout in each general election from 1984 to 2008. And from 2012 to 2020, I used <a href= "https://bniajfi.org/indicators/Sustainability/voted/2020" target="_blank" style="color: black;"> this site </a>, which has the voter turnout in Baltimore City from 2012 to 2020.

```{r include=FALSE}
#libraries and settings
library(tidyverse)
library(janitor)
library(formattable)
library(dplyr)
library(rvest)
library(rmarkdown)
library(knitr)

#turn off scientific notation
options(scipen=999)
```

```{r include=FALSE}
#upload csv and clean names/rows
pct_vote_csas <- read_csv("data/pct_vote_csa_clean.csv")

pct_vote_csas <- pct_vote_csas%>%
  rename(csa = 1, pct_voters = 2)

pct_vote_csas$csa[43] <- "Westport/Mount Winans/Lakeland" 
```

```{r}
#webscrape data table from prison policy initiative website 
incarceration_x2020 <- "https://www.prisonpolicy.org/origin/md/2020/baltimore_csa.html"
incarceration_x2020 <- incarceration_x2020 %>% 
  read_html() %>% 
  html_table()
incarceration_x2020 <- incarceration_x2020[[1]]%>%
  clean_names()%>%
  rename(csa = 1, rate_per_x100000 = 5)%>%
  select(csa, rate_per_x100000)
```

```{r include=FALSE}
baltcity_income <- read_csv("data/baltcity_income.csv")%>%
  clean_names()%>%
  rename(csa = community, median_income = x2020_data)%>%
  filter(csa != "Baltimore City")
```

```{r include=FALSE}
biz <- read_csv("data/smallbiz_csa.csv")

biz <- biz%>%
  clean_names()%>%
  rename(csa = community, num_biz_x1000 = x2020_data)%>%
  filter(csa != "Baltimore City")
```

```{r include=FALSE}
internet <- read_csv("data/internet.csv")

internet <- internet%>%
  clean_names()%>%
  rename(csa = community, pct_no_internet = x2020_data)%>%
  filter(csa != "Baltimore City")
```

```{r include=FALSE}
#voter turnout in the 2010 election
voting_2010 <- read_csv("data/voting_2010.csv")%>%
  clean_names()%>%
  rename(csa = community, pct_voters = x2010_data)%>%
  filter(csa != "Baltimore City")
```

```{r include=FALSE}
#voter turnout in the 2012 election
voting_2012 <- read_csv("data/voting_2012_clean.csv")%>%
  clean_names()%>%
  rename(csa = community, pct_voters = x2012_data)%>%
  filter(csa != "Baltimore City")
```

```{r include=FALSE}
#voter turnout in the 2014 election
voting_2014 <- read_csv("data/voting_2014.csv")%>%
  clean_names()%>%
  rename(csa = community, pct_voters = x2014_data)%>%
  filter(csa != "Baltimore City")
```

```{r include=FALSE}
#voter turnout in the 2016 election
voting_2016 <- read_csv("data/voting_2016.csv")%>%
  clean_names()%>%
  rename(csa = community, pct_voters = x2016_data)%>%
  filter(csa != "Baltimore City")
```

```{r include=FALSE}
#voter turnout in the 2018 election
voting_2018 <- read_csv("data/voting_2018.csv")%>%
  clean_names()%>%
  rename(csa = community, pct_voters = x2018_data)%>%
  filter(csa != "Baltimore City")
```

```{r include=FALSE}
#voter turnout in the 2022 election
voting_2016 <- read_csv("data/voting_2016.csv")%>%
  clean_names()%>%
  rename(csa = community, pct_voters = x2016_data)%>%
  filter(csa != "Baltimore City")
```

```{r include=FALSE}
trends_turnout <- read_csv("data/trends_turnout.csv")

trends_turnout <- trends_turnout%>%
  rename(pct_turnout = 2)
```

```{r include=FALSE}
#upload data
maryland_voting <- read_csv("data/maryland_voter_turnout_historic.csv")

#change this column name to make join easier later
maryland_voting <- maryland_voting%>%
  rename(state = 1)

#make two tables into separate datasets and rejoin
placeholder1 <- maryland_voting%>%
  slice(7:57)%>%
  rename("2020" =2, "2016" = 4, "2012" = 6, "2008" = 8, "2004" = 10)

placeholder2 <- maryland_voting%>%
  slice(75:125)%>%
  rename("2000" = 2, "1996" =4, "1992" = 6, "1988" = 8, "1984" = 10)

#join to make cleaner table - I'm going to use the total reported voters instead of just the number of citizens who were reported voting because that's what the Bcity table had it by
maryland_voting_clean <- inner_join(placeholder1, placeholder2, by = "state")%>%
  select("2020", "2016", "2012", "2008", "2004", "2000", "1996", "1992", "1988", "1984", state)

#filter to just maryland
maryland_voting_clean <- maryland_voting_clean%>%
  filter(state == "Maryland")%>%
  select(-state)

#switch rows with columns
maryland_cleanest_voting <- t(maryland_voting_clean)

#rename columns
colnames(maryland_cleanest_voting) <- c("pct_turnout")
```

```{r include=FALSE}
#upload csv 
precinct_voters_x2020 <- read_csv("data/cleaned_precinct_x2020.csv")

#remove rows with NAs and get rid of extra column copied from PDF
precinct_voters_x2020 <- precinct_voters_x2020 %>%
  filter(complete.cases(.))%>%
  slice(-111)%>%
  clean_names()%>%
  select(precinct, dem, rep, total)
```

```{r include=FALSE}
#voting by demographic 2020
demo_voting <- read_csv("data/demo_voting_x2020_finalll.csv", locale=locale(encoding="latin1"))%>%
  clean_names
#delete top seven rows for worcester copied from the page before and column three filled with only N/As
demo_voting <- demo_voting[-c(1:7), -3]%>%
  slice(-1)%>%
  rename(precinct = voting_district_vtd, precinct_pop = total_population, black_aa = black_or_african_american_alone)%>%
  select(precinct, precinct_pop, black_aa)
```

**Q1: Which five Baltimore City community statistical areas had the lowest voter turnout percentage in the 2020 general election? Which five had the highest?**

Lowest voter turnout percent

```{r}
#use slicemax/min to make new tables with top and bottom five csas by voter turnout
pct_vote_csas%>%
  slice_min(pct_voters, n = 5)

paged_table(pct_vote_csas)
```

Highest voter turnout percent

```{r}
pct_vote_csas%>%
  slice_max(pct_voters, n = 5)

paged_table(pct_vote_csas)
```

I created a custom map in flourish by uploading a GeoJSON file with the Baltimore City CSA boundaries, which I found through <a href = "https://data.baltimorecity.gov/datasets/9c96ae20e6cc41258015c2fd288716c4/explore" target="_blank" style="color: black;"> Open Baltimore </a>

```{r, results='asis'}
cat('<div class="flourish-embed flourish-map" data-src="visualisation/17796037"><script src="https://public.flourish.studio/resources/embed.js"></script></div>')
```

It's clear the areas with the highest voter turnout are in the middle areas of Baltimore, especially in the Inner Harbor area stretching north to Homeland and Guilford, with East and West Baltimore showing significantly less voter turnout. Sixteen CSAs had a voter turnout of less than 50%, and were all concentrated in similar areas.

**Q2: Are there correlations between the Baltimore City 2020 general election voter turnout and factors including incarceration rate, business activity, wealth, internet access and race?**

Let's compare all these factors in the five CSAs with the highest and lowest voter turnout rates

I'll start with incarceration rate and voter turnout

```{r}
#join incarceration_x2020 and pct_vote_csas datasets
incarceration_voting <- inner_join(incarceration_x2020, pct_vote_csas, by = "csa")

incarceration_voting %>%
  slice_min(pct_voters, n = 5)

incarceration_voting %>%
  slice_max(pct_voters, n = 5)
```

```{r, results='asis'}
cat('<div class="flourish-embed flourish-map" data-src="visualisation/17809409">
      <script src="https://public.flourish.studio/resources/embed.js"></script>
    </div>')
```

This visualization shows there is a correlation between incarceration rate and voter turnout. Looking at the map of voter turnout next to the map of incarceration rates, it's clear that the two are generally inverses of on another, with areas in North and Central Baltimore having a low incarceration rate and high voter turnout and East and West Baltimore having a higher incarceration rate and low voter turnout.

```{r, results='asis'}
cat('<div class="flourish-embed flourish-scatter" data-src="visualisation/17795149"><script src="https://public.flourish.studio/resources/embed.js"></script></div>')
```

There is a negative relationship between incarceration rate and voter turnout. As the rate of incarceration per 100,000 people increases, the percent of registered voters decreases.

Just to be really sure, let's compute some correlation coefficients. Values close to 1 or -1 indicate a strong positive or negative correlation, while values closer to zero indicate a weak correlation.

Correlation between incarceration rate and voter turnout.

```{r}
#convert all values to numeric for calculation
incarceration_voting$rate_per_x100000 <- gsub("[, ]", "", incarceration_voting$rate_per_x100000)
incarceration_voting$rate_per_x100000 <- as.numeric(incarceration_voting$rate_per_x100000)

#compute correlation coefficient
cc_incarceration <- cor(incarceration_voting$pct_voters, incarceration_voting$rate_per_x100000, method = "pearson", use = "complete.obs")

cc_incarceration
```

A correlation coefficient of -0.82 indicates a strong negative correlation, meaning areas of lower incarceration rates likely have higher voter turnout rates.

```{r}
#join baltcity_income and pct_vote_csas
wealth_voting <- inner_join(pct_vote_csas, baltcity_income, by = "csa")

wealth_voting_min <- wealth_voting%>%
  slice_min(pct_voters, n = 5)

wealth_voting_max <- wealth_voting%>%
  slice_max(pct_voters, n = 5)

paged_table(wealth_voting_min)

paged_table(wealth_voting_max)
```

```{r, results='asis'}
cat('<div class="flourish-embed flourish-map" data-src="visualisation/17824036">
      <script src="https://public.flourish.studio/resources/embed.js"></script>
    </div>')
```

Areas of higher median household income are concentrated in the middle of Baltimore City and in the northern areas. This is a similar pattern to the voter turnout rate map.

```{r, results='asis'}
cat('<div class="flourish-embed flourish-scatter" data-src="visualisation/17824265"><script src="https://public.flourish.studio/resources/embed.js"></script></div>')
```

There is a positive relationship between median household income and voter turnout. As the median household income increases, the percent of registered voters decreases.

Let's also calculate the correlation coefficient between median household income and voter turnout

```{r}
cc_wealth <- cor(wealth_voting$pct_voters, wealth_voting$median_income, method = "pearson", use = "complete.obs")

cc_wealth
```

A correlation coefficient of 0.81 indicates a strong positive correlation between voter turnout and median household income.

Number of businesses per 1,000 people and voter turnout

```{r}
#join biz and pct_vote_csas dataset
biz_voting <- inner_join(biz, pct_vote_csas, by = "csa")

biz_voting_min <- biz_voting %>%
  slice_min(pct_voters, n = 5)

biz_voting_max <- biz_voting%>%
  slice_max(pct_voters, n = 5)

paged_table(biz_voting_min)

paged_table(biz_voting_max)
```

There doesn't seem to be a significant correlation between number of businesses per 1,000 residents and voter turnout. There are quite a few outliers, including Oldtown/Middle East, which was one of the lowest voter turnout areas, but had the second highest amount of businesses per 1,000 residents. This could be for a variety of reasons. For one, many of the areas that have high voter turnout are in wealthier, suburban areas with less concentration of businesses.

Correlation between number of businesses per 1,000 residents and voter turnout

```{r}
#compute correlation coefficient
cc_biz <- cor(biz_voting$pct_voters, biz_voting$num_biz_x1000, method = "pearson", use = "complete.obs")

cc_biz
```

There is little to no correlation between the amount of businesses and voter turnout, as indicated by a correlation coefficient of -0.02.

Percent of households without internet and voter turnout

```{r}
#join internet and pct_vote_csas dataset
internet_voting <- inner_join(internet, pct_vote_csas, by = "csa")

internet_voting_min <- internet_voting %>%
  slice_min(pct_voters, n = 5)

internet_voting_max <- internet_voting %>%
  slice_max(pct_voters, n = 5)

paged_table(internet_voting_max)

paged_table(internet_voting_min)
```

```{r, results='asis'}
cat('<div class="flourish-embed flourish-scatter" data-src="visualisation/17824515"><script src="https://public.flourish.studio/resources/embed.js"></script></div>')
```

There is a negative relationship between percent of households without internet access and voter turnout. As the percent of households without internet increases, the percent of registered voters decreases.

Correlation between percent of households without internet access and voter turnout

```{r}
cc_internet <- cor(internet_voting$pct_voters, internet_voting$pct_no_internet, method = "pearson", use = "complete.obs")

cc_internet
```

A correlation coefficient of -0.71 indicates a strong negative correlation between voter turnout and median household income

**Q3: How has voter turnout in Baltimore City changed from 2010 to 2020? Which CSA had the biggest difference between the 2012 and 2020 general election?**

Let's look at community statistical areas first

```{r}
#rename row from "&" to "and" to match up with the rest of the voting data used later
pct_vote_csas$csa <- gsub("&", "and", pct_vote_csas$csa)

#make one dataset for combined 2012 and 2020
voting_x2012_x2020 <- inner_join(voting_2012, pct_vote_csas, by = "csa")

voting_x2012_x2020 <- voting_x2012_x2020 %>%
  rename(x2012_turnout = 2, x2020_turnout = 3) %>%
  mutate(pct_change = ((x2020_turnout - x2012_turnout) / x2012_turnout) * 100)

voting_x2012_x2020_max <- voting_x2012_x2020 %>%
  slice_max(pct_change)

voting_x2012_x2020_min <- voting_x2012_x2020 %>%
  slice_min(pct_change)
```

Orangeville/East Highlandtown had the biggest change, with an almost 65% increase in voter turnout.

Clifton-Berea had the largest decrease in voter turnout, dropping from 57.5% in 2012 to 46.4% in 2020.

Let's now look at how voter turnout in community statistical areas has changed over time. To do this, I'll first make a combined dataset and then visualize it in Flourish.

```{r message=FALSE, warning=FALSE}
voting_all <- inner_join(voting_2010, voting_2012, by = "csa") %>%
                 inner_join(voting_2014, by = "csa") %>%
                 inner_join(voting_2016, by = "csa") %>%
                 inner_join(voting_2018, by = "csa") %>%
                 inner_join(pct_vote_csas, by = "csa")

voting_all <- voting_all%>%
  rename(x2010_turnout = 2, x2012_turnout = 3, x2014_turnout = 4, x2016_tunout = 5, x2018_turnout = 6, x2020_turnout = 7)
```

Press play to see change overtime between gubernational and general elections in the last 10 years

```{r, results='asis'}
cat('<div class="flourish-embed flourish-map" data-src="visualisation/17832534"><script src="https://public.flourish.studio/resources/embed.js"></script></div>')
```

Gubernatorial elections tend to get less voter turnout than general elections, which is unsurprising as that is a <a href="https://www.baltimoresun.com/2022/12/22/only-half-of-marylands-voters-cast-ballots-in-2022-a-historically-low-number-for-a-midterm-election/" target="_blank" style="color: black;">well-documented phenomenon</a>. But it seems that 2020 has the highest voter turnout out of the 10-year period. It also seems that the "Black Butterfly White L" phenomenon becomes more pronounced over time

Let's get a deeper understanding of voter turnout by looking at Baltimore City as a whole and comparing it to the state of Maryland. I'm also going to look at a wider range of years and focus on general election turnout because it's such a wide spread of data.

```{r, results='asis'}
cat('<div class="flourish-embed flourish-chart" data-src="visualisation/17832645"><script src="https://public.flourish.studio/resources/embed.js"></script></div>')
```

Looking at a wider spread of data shows that while voter turnout was highest in 2020 over the past 12 years, voter turnout significantly decreased in the 2008 through 2023 period. 68% the city's voting-age population voted in the 2008 general election compared to 58.8% in 2020's general election.

**Q4: In August 2020, which parts of the city had the most registered Democrats? Which had the most Republicans?** Community statistical areas are great for data analysis because they're more recognizable to readers, but precinct-level data leads to more detailed results. While each voting precinct has an address with a voting location, the precinct is an area, so it would be inaccurate to assign each to a community statistical area or neighborhood because there would be some precincts that are partially in one and partially in another. That's why maps are especially helpful when working with this data

```{r}
#add new column for percent of whole
precinct_voters_x2020 <- precinct_voters_x2020%>%
  mutate(pct_dem = (dem / total)*100)%>%
  mutate(pct_rep = (rep / total)*100)

precinct_voters_x2020_min <- precinct_voters_x2020%>%
  slice_min(pct_dem)

precinct_voters_x2020_max <- precinct_voters_x2020%>%
  slice_max(pct_rep)

paged_table(precinct_voters_x2020_min)

paged_table(precinct_voters_x2020_max)
```

The precinct with the lowest percent of Democrats still had 53% in August 2020, which means the majority of the city are Democrats.

The precinct with the highest percent of Republicans only had 23.7% in August 2020

Let's see if there's a correlation with the amount of Black residents who are over the voting age of 18

```{r message=FALSE, warning=FALSE}
#add a column for percent Black residents of whole population in precinct
demo_voting <- demo_voting%>%
  mutate(pct_black = (black_aa / precinct_pop)*100)
```

Voting precincts don't make much sense without a visualization. I uploaded the geojson for 2020 voting precincts on the Open Baltimore site I got the first geojson. This map has a toggle which allows users to switch from concentration of Black residents over the age of 18 by precinct and concentration of Democrats

```{r, results='asis'}
cat('<div class="flourish-embed flourish-scatter" data-src="visualisation/17824265" style="width: 100%; height: 600px;">
      <script src="https://public.flourish.studio/resources/embed.js"></script>
    </div>')
```

There were higher concentrations of registered Democrats in East and West Baltimore City in August 2020 than in the North or South. The clusters of precincts with the highest concentration of Democrats are in neighborhoods such as Upton/Druid Heights, Oldtown/Middle East and Westport/Mount Winans/Lakeland. There was also a significantly higher concentration of Black residents over the age of 18 in the same East and West Baltimore Areas.

Let's look at Republicans now

```{r, results='asis'}
cat('<div class="flourish-embed flourish-map" data-src="visualisation/17835183"><script src="https://public.flourish.studio/resources/embed.js"></script></div>')
```

There are not many Republicans in Baltimore City. Those that are reside in the middle and more rural parts of the city. Note: sometimes this graphic doesn't load right away on the page so you might have to reload and it should pop up.
