#Lab_03 {lab_03}

rob wells, 8/14/2021

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse here
library(tidyverse)

# Load janitor here
library(janitor)
```

### Load data

In this data set, each row represents a city or town in Maryland with the 2010 and 2020 figures for: 

-  Total population 
-  White only population
-  Black or African American only population
-  Hispanic only population 
-  Native American only population
-  Asian only population
-  Pacific Islander (including native Hawiian) only population

Background on this data from [Social Explorer](https://www.socialexplorer.com/data/ACS2016_5yr/metadata/?ds=ACS16_5yr&table=B02001) 
```{r}
# Write code to load the Maryland race by city, place for 2010 - 2020 and call it md_race.
# The data is called lab_03.csv, and is in the same folder as lab_03.qmd (this data notebook)
# your answer here
md_race <- read_csv("/Users/oliviaborgula/Documents/GitHub/olivia_borgula_jour472/lab_03/lab_03.csv") %>% as.data.frame()
```


## Answer questions

Q1: Which five cities have the largest overall Hispanic population in 2020? Name the five cities that had the largest Hispanic population in 2020. 

A1:
```{r}

# Write your code in this codeblock.
md_race %>% 
  select(place, x2020_hispanic) %>% 
  slice_max(x2020_hispanic, n = 5)

# your answer here
#Baltimore City, Wheaton CDP, Germantown CDP, Silver Spring CDP and Aspen Hill CDP have the highest Hispanic population

```

Q2: Find the cities with the largest Black population and arrange the results in descending order. 

A2:

```{r}
#write your code here
md_race %>% 
  select(place, x2020_black) %>% 
  arrange(desc(x2020_black))

# your answer here
#The city with the largest Black population is Baltimore City. The next cities with the highest population of Black people are Waldorf CDP, Bowie City, Clinton CDP and Columbia CDP. 

```

Q3: Create a table called top_hispanic that shows the top 15 communities with the largest Hispanic population in 2020, and also display the 2010 Hispanic population and the total community population for 2020. Look at your results - what stands out to you? Craft a one-paragraph story pitch for your editor. 

A3:

```{r}

# your answer here
top_hispanic <- md_race %>% 
  select(place,x2020_hispanic,x2010_hispanic,x2020_total) %>% 
  slice_max(x2020_hispanic, n = 15)

#The 15 cities with the highest Hispanic population in 2020 all saw an increase in the number of Hispanic residents in the last decade. Seven out of those 15 cities are in Montgomery County. In Aspen Hill CDP, the Hispanic population increased by around 81% during that time, growing from 10,280 to 18,635. But the unincorporated city is still only 4th on the list, with Baltimore City maintaining the highest Hispanic population of 32,627 in 2020. I would explore what's causing these increases in the Hispanic population, why so many are in Montgomery County (is it just because it's the most populous county or another reason?) I would also look into how being a CDP impacts a city. There are 10 cities on this list that are CDPs, and they generally don't have city councils. Does that make Hispanic residents feel underrepresented? I would interview Hispanic residents in the area, preferably those who have lived there for 10 years or longer about whether they've noticed this increase. I would also speak with Hispanic activists about what the rise in the Hispanic population means for Hispanic people, Montgomery County executives (Marc Elrich?) and leaders from other mentioned cities such as city council members or mayors about whether they've implemented any policies or started any initiatives based on the changing demographics. 
```

