#lab_11
Rob Wells, April 11, 2024

## About this lab

To complete this lab, you need to:
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Overview

There are three main goals of this lab:

* learn to make use of R packages that can access data via third-party APIs.
* learn to develop a quick and accurate understanding of what data can be obtained via third-party APIs. 
* learn to develop questions that can be answered using that data, and avoid developing questions that cannot. 

## APIs to choose from
 

* tidycensus - for asking census-related questions.

## Load libraries

**Task**: load the tidyverse and janitor.  Uncomment the one other package you plan to work with.  Do not choose more than one API package.

```{r}
library(tidyverse)
library(janitor)
library(tidycensus)

```

## Get setup to access third-party APIs

### tidycensus

Use the U.S. census access token you created for this week's lab.  You can store it by editing the block below to replace 'YOUR KEY HERE' with your actual key, a long string of letters and numbers. If you need to sign up for another one, you can do so here:  [apply for an API key with the Census Bureau](https://api.census.gov/data/key_signup.html). 

```{r}
census_api_key("dfcb820a29176622bc44d5c66b77d03b5675937d", install=TRUE, overwrite = TRUE)
readRenviron("~/.Renviron")
```



### tidycensus

This package is designed to help you pull demographic information about different U.S. geographies.

* Basic usage https://walker-data.com/tidycensus/articles/basic-usage.html
* Functions: https://walker-data.com/tidycensus/reference/index.html
* More details https://walker-data.com/tidycensus/

Sample question: which state saw the largest growth in Hispanic population between 2010 and 2020?

```{r}

###
# Load ACS variables
###

acs_vars <- load_variables(dataset = "acs5", year = 2020)

###
# Get median household income for Maryland 2020
###

md <- get_acs(geography = "county",
              variables = c(median_income = "B19013_001"),
              state = "MD",
              year = 2020) 


```

### Develop and answer questions

**Task** Develop two different *interesting* questions that you can answer with the data. Run them by your professor before getting started on answering, to ensure they're sufficiently interesting and possible with the data. 


**Which state has had the largest increase in Spanish-speaking residents in the last 10 years? Which state hs the largest increase in Spanish-speaking residents that reported speaking English "less than very well"?**
```{r}
#get data from census for all Spanish-speaking residents in last 10 years
x2010_spanish <- get_acs(geography = "state",
              variables = c(total_spanish_2010 = "B16001_003"),
              year = 2010) 

x2020_spanish <- get_acs(geography = "state",
              variables = c(total_spanish_2020 = "B16001_003"),
              year = 2020) 

#join these columns 
total_spanish <- x2010_spanish %>%
  inner_join(x2020_spanish, by="GEOID")

#perform calculations
total_spanish <- total_spanish %>%
  mutate(pct_change = percent(estimate.y - estimate.x)/estimate.x)%>%
  arrange(desc(pct_change))

#North Dakota saw the largest increase in Spanish-speaking residents, with a 62.5% increase from 2010 to 2020. 8,183 residents were Spanish-speaking in 2010 compared to 13,295 in 2020.

#The above calculation is for all Spanish-speaking residents. I now want to look at which state had the highest increase in Spanish-speaking residents who reported speaking English "less than very well" from 2010 to 2020. 

#get data from census for all Spanish-speaking residents in last 10 years
x2010_engl_fluency <- get_acs(geography = "state",
              variables = c(engl_fluency_2010 = "B16001_005"),
              year = 2010) 

x2020_engl_fluency <- get_acs(geography = "state",
              variables = c(engl_fluency_2020 = "B16001_005"),
              year = 2020) 

#join these columns 
total_engl_fluency <- x2010_engl_fluency %>%
  inner_join(x2020_engl_fluency, by="GEOID")

#perform calculations
total_engl_fluency <- total_engl_fluency %>%
  mutate(pct_change = percent(estimate.y - estimate.x)/estimate.x)%>%
  arrange(desc(pct_change))

#Again, North Dakota was the state with the largest increase of 116.7%. 1787 people reported sepaking English "less than very well" in 2010 compared with 3873 in 2020.  

```

**Which state has had the largest increase in residents living below the poverty line in the last 10 years ? **

```{r}
#get data from census for poverty status of US residents in last 10 years. This data is telling us how many people in the given area have been determined to be poverty status
x2010_poverty_status <- get_acs(geography = "state",
              variables = c(poverty_status_2010 = "B06012_001"),
              year = 2010) 

x2020_poverty_status <- get_acs(geography = "state",
              variables = c(poverty_status_2020 = "B06012_001"),
              year = 2020)

#join these columns 
total_pv_status <- x2010_poverty_status %>%
  inner_join(x2020_poverty_status, by="GEOID")

#perform calculations - I'm not using formattable here because that converts the values into characters and makes it difficult to visualize
total_pv_status <- total_pv_status %>%
  mutate(pct_change = (estimate.y - estimate.x) / estimate.x * 100)%>%
  arrange(desc(pct_change))

#The District of Columbia has seen the largest increase in residents living below the poverty line, with a 21.4% rise from 2010 to 2020. There were 551,331 people living in poverty in 2010 compared to 669,089 living in poverty in 2020.

```


**Task** Write code to answer the questions.  Write up a sentence with each of your findings. Produce at least one visualization to support one of your findings.

**Sentence with each of my findings**
```{r}
#North Dakota saw the largest increase in Spanish-speaking residents out of all US states, with a 62.5% increase from 2010 to 2020 as well as a 116.7% increase in Spanish-speakng residents who reported speaking English "less than very well" in the same time period. 

#The District of Columbia has seen the largest increase in residents living below the poverty line out of all US states, with a 21.4% rise from 2010 to 2020. 

```

**Data visualization**
```{r}
#make new dataset with just the top 10 states with highest increase in residents in poverty from 2010 to 2020
top_10_pv_status <- total_pv_status %>%
  slice_head(n = 10)

#visualize it 
ggplot(top_10_pv_status, aes(x = reorder(NAME.x, pct_change), y = pct_change)) +
  geom_bar(stat = "identity", fill = "darkblue", color = "black") +
  labs(
    title = "Top 10 States with Largest Increase in Residents in Poverty from 2010 to 2020",
    x = "State",
    y = "Percentage Change",
    caption = "Data source: American Community Survey"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip()
```


