
# Math - Percentage of Whole Calculations 

Here are a few refresher exercises for doing math calculations within the code.


```{r}
library(tidyverse)
library(formattable)
```

We'll import a dataset of median household income in Baltimore for the 2010, 2016 and 2020 surveys. The data is in the data folder in this chapter's pre-lab directory. We'll use this to explore ways to create new information from existing data.

/Users/robwells/Code/data_journalism_class/04_labs/lab_05/pre_lab_05/baltcity_income_clean.csv

```{r}

baltcity_income<- read_csv("https://raw.githubusercontent.com/wellsdata/data_journalism_class/main/04_labs/lab_05/pre_lab_05/baltcity_income_clean.csv") %>% 
  as.data.frame()


```
Let's see which neighborhoods are the largest on a percentage basis in 2020 

```{r}
#add column showing neighborhoods' populations as percent of a whole
baltcity_income <- baltcity_income %>%
  mutate(pct_whole = percent(x2020 / sum(x2020, na.rm = TRUE)))%>%
  arrange(desc(pct_whole))
  
```

Create a Diff_Income column and then a percentage change column for 2020 vs 2010

```{r}
baltcity_income <- baltcity_income%>%
  mutate(Diff_Income = percent(x2020 - x2010)/x2010)
  
```
