```{r}
#turn off scientific notation
options(scipen=999)

```

---
names: Olivia Borgula and Lizzy Alspach
date: 4/4/2024
assignment: reverse engineering project milestone one
---

**INTRODUCTION** This project will reverse engineer findings by The Washington Post's story, headlined "Nearly 250 women fatally shot by police since 2015." The link is here: https://www.washingtonpost.com/graphics/2020/investigations/police-shootings-women/

This story describes the disparity in people fatally shot by police in regards to sex and within sex, race. We will replicate the following findings described in paragraphs in The Post’s story:

Q1: “Since The Washington Post began tracking fatal shootings by police in 2015, officers have fatally shot 247 women out of the more than 5,600 people killed overall.”

Q2: “Of the 247 women fatally shot, 48 were Black and seven of those were unarmed. At least 89 of the women were at their homes or residences where they sometimes stayed. And 12 of those women killed at home were shot by officers who were there to conduct a search or make an arrest.”

Q3: “Since 2015, Black women have accounted for less than 1 percent of the overall fatal shootings in cases where race was known. But within this small subset, Black women, who are 13 percent of the female population, account for 20 percent of the women shot and killed and 28 percent of the unarmed deaths.”

Q4: “About 31 percent, or 77, of the 247 women fatally shot by police since 2015 had mental health issues, compared to 22 percent of the 5,362 men killed.”

**SETUP** Load libraries and turn off scientific notation

```{r}
#load libraries
library(tidyverse)
library(janitor)
library(formattable)
library(dplyr)
```

Upload data

```{r}
shootings <- read_csv("data/fatal-police-shootings.csv")%>%
  clean_names()

agencies <- read_csv("data/fatal-police-shootings-agencies.csv")%>%
  clean_names()
```

Filter out data past Sept. 4, 2020, when the article was published

```{r}
#filter for specific times that happened before sept. 4, 2020, when the story was published
filtered_shootings <- shootings %>%
  filter(date <= as.Date("2020-09-04"))

count(filtered_shootings)
#shows next part of the question, with more than 5,600 deaths
```

**Q1: Since The Washington Post began tracking fatal shootings by police in 2015, officers have fatally shot 247 women out of the more than 5,600 people killed overall**

```{r}
#filter for women
women_shootings <- filtered_shootings%>%
  filter(gender == "female")

#count total women shot
count(women_shootings)

#there were 248 women killed out of the more than 5,600 overall. The disparity in our results vs. the Washington Post's analysis is due to updates to the database in the last four years. Likely, more information was released about a women who was shot after publication, which changed the database 
```

**Q2: Of the 247 women fatally shot, 48 were Black and seven of those were unarmed. At least 89 of the women were at their homes or residences where they sometimes stayed. And 12 of those women killed at home were shot by officers who were there to conduct a search or make an arrest.**

```{r}
#filter by race
race_women_shootings <- women_shootings %>% 
  filter(race == "B")

count(race_women_shootings)

#48 of the women shot were Black 

#out of the black women, find how many were unarmed
black_women_unarmed <- race_women_shootings %>% 
  filter(armed_with == "unarmed")

count(black_women_unarmed)

#7 Black women who were shot were unarmed
```

We don't have sufficient data accessible to us to replicate the findings about women being at home and officers conducting home searches:

"At least 89 of the women were at their homes or residences where they sometimes stayed. And 12 of those women killed at home were shot by officers who were there to conduct a search or make an arrest."

While the database contains latitude and longitude coordinates about where the women were killed, the Washington Post GitHub for this database does not contain any information or data about the women's home addresses. 

**Q3: Since 2015, Black women have accounted for less than 1 percent of the overall fatal shootings in cases where race was known. But within this small subset, Black women, who are 13 percent of the female population, account for 20 percent of the women shot and killed and 28 percent of the unarmed deaths**

first part of graf: Black women have accounted for less than 1 percent of the overall fatal shootings in cases where race was known

```{r}
#get datasets for Black women who were fatally shot and total number of people who were fatally shot and exclude rows that have an "NA" value for race ("...where race was known")
total_rows_bw <- sum(!is.na(race_women_shootings$race))
total_rows_shootings <- sum(!is.na(filtered_shootings$race))

#calculate the percent 
percent(total_rows_bw / total_rows_shootings)


#Black women account for just under than 1 percent of the overall fatal shootings, which is slightly off due to updates in the dataset 
```

second part of graf: Black women account for 20 percent of the women shot and killed and 28 percent of the unarmed deaths

```{r}
#find number of rows in dataframe for Black women and female shootings dataframe with rows removed that contain "N/A" for race (assuming "where race was known" also applies to these sentences)
total_rows_black_women <- sum(!is.na(race_women_shootings$race))
total_rows_female <- sum(!is.na(women_shootings$race))

#calculate the percent
percent(total_rows_black_women / total_rows_female)

#Black women account for a little more than 19 percent of the women shot, or close to 20 percent - this figure is slightly off likely due to updates in the dataframe in the four years since the article was published. 
```

third part of graf: Black women are 13 percent of the female population

```{r}
#for this part, reporters likely did not do the same level of data analysis as other sections since basic Census Bureau data like this is easily accessible online. They could find the information through this page, which shows the number of females in the US population. Under "Race and Hispanic Origin", it lists Black or African American as about 13%: https://www.census.gov/quickfacts/fact/table/US/LFE046222 - they likely would have used a slightly more specific table showing only 2020 data, but this is the closest we can find in 2024. 
```

fourth part of graf: Black women account for 28 percent of the unarmed deaths

```{r}
#filter for unarmed females
female_unarmed <- filtered_shootings %>%
  filter(armed_with == "unarmed", gender == "female")

#Find number of rows excluding those with "N/A" for race (black_women_unarmed doesn't have any "N/A" values for race, but adding based on mindset that wapo reporters are working with regularly updated data)
total_rows_black_women_unarmed <- sum(!is.na(black_women_unarmed$race))
total_rows_female_unarmed <- sum(!is.na(female_unarmed$race))

#find percent change
percent(total_rows_black_women_unarmed / total_rows_female_unarmed)

#Black women accounted for close to 28% of unarmed deaths. This isn't the exact figure likely due to a database update since the time of publication.
```

**Q4: About 31 percent, or 77, of the 247 women fatally shot by police since 2015 had mental health issues, compared to 22 percent of the 5,362 men killed.**

first part of graf: about 31 percent, or 77, of the 247 women fatally shot by police since 2015 had mental health issues
```{r}
mental_health_women <- women_shootings %>% 
  filter(was_mental_illness_related == "TRUE")

count(mental_health_women)

#83 women had mental health issues 

total_women_mental_health <- sum(!is.na(mental_health_women))
total_women_shot <- sum(!is.na(women_shootings))

percent(total_women_mental_health / total_women_shot)

#This is slightly more than what it was previously, seeing as there are now 83 women recorded with mental health issues and 248 women recorded fatalities. Now, it's at just over 33 percent.

#Since this database has been continuously updated since the publication of this investigation in 2020, there are likely to have been about six more women who were discovered to have mental health issues. 
```


second part of graf: second part of graf: 22 percent of the 5,362 men killed had mental health issues
```{r}
#make a dataframe just for men
men_shot <- filtered_shootings %>% 
  filter(gender == "male")

count(men_shot)
#this count will also be off, because from this time it appears The Post discovered that there were more men who were murdered in the timeframe but only discovered after this investigation was published

#make a dataframe for men with mental health issues
men_mental_health <- men_shot %>% 
  filter(was_mental_illness_related == "TRUE")

#calculate the percent total
total_men_mental_health <- sum(!is.na(men_mental_health))
men_shot_total <- sum(!is.na(men_shot))

percent(total_men_mental_health / men_shot_total)

#this total is also a little off, seeing as there have been more men discovered to have mental health issues since this investigation has been published, so we get that just over 23% had mental health issues.

```

**MEMO**
Out of the nearly 250 women who have been fatally shot by police from 2015 to 2020, Black women and females with mental health issues were killed at a disproportionately high rate. 

While Black women are 13 percent of the female population, they account for 20 percent of the women shot and killed and 28 percent of the unarmed deaths. 

About 31 percent of women fatally shot had mental health issues, compared to 22 percent of the 5,362 men killed from 2015 to 2020. More than a third of all women who were fatally shot were killed in their homes or residences where they sometimes stayed. 

The reporters brought this story to life primarily through graphic presentation and thorough storytelling. As soon as users open the story, the interactivity of the story’s site allows readers to scroll through a graphic representing the number of women who were fatally shot through dots. Large blocks of texts are separated by captivating photography and graphs breaking down demographic information. 

The story is also rooted in rich reporting, with reporters interviewing families of people who police fatally shot, criminal justice experts, sociologists and anti-violence activists. Reporters don’t treat the women in the stories as statistics, they write about who they were before they died and their impact on the community. 

The reporters approached the story by using data to give a comprehensive overview of women who were fatally shot by police, a topic that previously had not been frequently quantified in mainstream news outlets. This was effective because they used a database to get a baseline understanding of how many women were shot, then went further and examined how factors such as race and mental health were represented in the data. 

They communicated these findings in simple language and did not bog down paragraphs with figures. Reporters frequently used modifiers such as “about,” “nearly” and “just over” to make data more digestible to readers. Not only that, but reporters did not solely use statistics from their data analysis. As discussed in many journalism courses, it’s best to find digestible ways to describe numbers and findings, such as “more than doubled.” These reporters did a nice job of ensuring figures were understandable, and they also complemented their figures with graphics, which made it even more comprehensive. 

Reporters could have improved this by adding even more interactive features and graphs. While the story already uses a lot of these techniques, I liked how stories such as the AP’s lethal force investigation allowed users to interact more directly with graphs within the article rather than just at the beginning of the story. 

Reporters specified exactly how they performed data analysis whenever possible. For figures about Black women being killed at a higher rate than other races, they clarify that the figures are only when data for race is present. They made it easy for readers to see their work by linking their public GitHub page and police shootings database. 

There’s also a methodology section at the bottom of the article that gives a general overview of how data was analyzed. It specifies that Census data is from five-year ACS estimates and that The Post labeled people as the gender they identified as even if it differs from their biological sex. It also links to a more comprehensive methodology report on another page. 

The datasets used for this piece are a part of The Post’s larger police shootings database, which contains updated data about all people killed by police since 2015. Data journalists did extensive work to prepare and clean this dataset. They obtained data by manually culling local news reports, collecting information from social media and law enforcement websites and monitoring independent databases such as Fatal Encounters and the Killed by Police project. 

Reporters then had to compile vast amounts of information from various sources, which they did manually and by using OCRs and other tools to extract information from PDFs. They then had to standardize the data by cleaning column names and manually inserting data into the correct place. 

They also had to develop tools that would efficiently update the database when reporters uncovered new information about the person or the shooting. But the frequent updates made it more difficult for us to replicate the reporter’s findings, because updated information led to  discrepancies between our figures and the ones in the article.

And while language such as “about” and “nearly” makes the story more digestible for readers, it was more difficult from a data journalists’ perspective because we didn’t have exact figures to compare our results to.  


