---
names: Olivia Borgula and Lizzy Alspach
date: 4/4/2024
assignment: reverse engineering project milestone one
---

**INTRODUCTION** This project will reverse engineer findings by The Washington Post's story, headlined "Nearly 250 women fatally shot by police since 2015." The link is here: https://www.washingtonpost.com/graphics/2020/investigations/police-shootings-women/

This story describes the disparity in people fatally shot by police in regards to sex and within sex, race. We will replicate the following findings described in paragraphs in The Post’s story:

Q1: “Since The Washington Post began tracking fatal shootings by police in 2015, officers have fatally shot 247 women out of the more than 5,600 people killed overall.”

Q2: “Of the 247 women fatally shot, 48 were Black and seven of those were unarmed. At least 89 of the women were at their homes or residences where they sometimes stayed. And 12 of those women killed at home were shot by officers who were there to conduct a search or make an arrest.”

Q3: “Since 2015, Black women have accounted for less than 1 percent of the overall fatal shootings in cases where race was known. But within this small subset, Black women, who are 13 percent of the female population, account for 20 percent of the women shot and killed and 28 percent of the unarmed deaths.”

Q4: “About 31 percent, or 77, of the 247 women fatally shot by police since 2015 had mental health issues, compared to 22 percent of the 5,362 men killed.”

**SETUP** Load libraries and turn off scientific notation

```{r}
#load libraries
library(tidyverse)
library(janitor)
library(formattable)
library(dplyr)

#turn off scientific notation
options(scipen=999)
```

Upload data

```{r}
shootings <- read_csv("data/fatal-police-shootings.csv")%>%
  clean_names()

agencies <- read_csv("data/fatal-police-shootings-agencies.csv")%>%
  clean_names()
```

Filter out data past Sept. 4, 2020, when the article was published

```{r}
#filter for specific times that happened before sept. 4, 2020, when the story was published
filtered_shootings <- shootings %>%
  filter(date <= as.Date("2020-09-04"))

count(filtered_shootings)
#shows next part of the question, with more than 5,600 deaths
```

**Q1: Since The Washington Post began tracking fatal shootings by police in 2015, officers have fatally shot 247 women out of the more than 5,600 people killed overall**

```{r}
#filter for women
women_shootings <- filtered_shootings%>%
  filter(gender == "female")

#add up total number of women shot
nrow(women_shootings)
#248 women

#add up total number of people shot
nrow(filtered_shootings)
#5,614 shootings

#there were 248 women killed out of the more than 5,600 overall. The disparity in our results vs. the Washington Post's analysis is due to updates to the database in the last four years. Likely, more information was released about a women who was shot after publication, which changed the database 
```

**Q2: Of the 247 women fatally shot, 48 were Black and seven of those were unarmed. At least 89 of the women were at their homes or residences where they sometimes stayed. And 12 of those women killed at home were shot by officers who were there to conduct a search or make an arrest.**

```{r}
#find number of women shot
count(women_shootings)

#248 women were fatally shot (see above note about discrepancies)

#filter to Black women and count number
black_women <- women_shootings %>% 
  filter(race == "B")
count(black_women)

#48 of the women shot were Black 

#filter to show only unarmed Black women 
black_women_unarmed <- black_women %>% 
  filter(armed_with == "unarmed")

count(black_women_unarmed)

#7 Black women who were shot were unarmed
```

The findings about women being at home/offices conducting home searches are impossible to replicate, seeing as the dataframes do not have columns specifically about the location of the victim and their addresses are not released, likely due to confidentiality reasons

**Since 2015, Black women have accounted for less than 1 percent of the overall fatal shootings in cases where race was known. But within this small subset, Black women, who are 13 percent of the female population, account for 20 percent of the women shot and killed and 28 percent of the unarmed deaths**

first part of graf: Black women have accounted for less than 1 percent of the overall fatal shootings in cases where race was known\*\*

```{r}
#get datasets for Black women who were fatally shot and total number of people who were fatally shot and exclude rows that have an "NA" value for race ("...where race was known")
total_rows_black_women <- sum(!is.na(black_women$race))
total_rows_shootings <- sum(!is.na(filtered_shootings$race))

#calculate the percent 
percent(total_rows_black_women / total_rows_shootings)

#Black women account for 0.86% of the overall fatal shootings in cases where race was known, or less than 1 percent. We don't know if this too far from 1%, but is the value we received.
```

second part of graf: Black women account for 20 percent of the women shot and killed and 28 percent of the unarmed deaths

```{r}
#find number of rows in dataframe for Black women and female shootings dataframe with rows removed that contain "N/A" for race (assuming "where race was known" also applies to these sentences)
total_rows_black_women <- sum(!is.na(black_women$race))
total_rows_female <- sum(!is.na(women_shootings$race))

#calculate the percent
percent(total_rows_black_women / total_rows_female)

#Black women account for 19.35% of the women shot, or close to 20 percent - this figure is slightly off likely due to updates in the dataframe in the four years since the article was published. We're unsure as to why the number is lower, but it's unclear if the number changed since the investigation was published.
```

third part of graf: Black women are 13 percent of the female population

```{r}
#for this part, reporters likely did not do the same level of data analysis as other sections since basic Census Bureau data like this is easily accessible online. They could find the information through this page: https://www.census.gov/data/tables/2023/dec/2020-census-demographic-profile.html#data
```

fourth part of graf: Black women account for 28 percent of the unarmed deaths

```{r}
#filter for unarmed females
female_unarmed <- filtered_shootings %>%
  filter(armed_with == "unarmed", gender == "female")

#Find number of rows excluding those with "N/A" for race (black_women_unarmed doesn't have any "N/A" values for race, but adding based on mindset that wapo reporters are working with regularly updated data)
total_rows_black_women_unarmed <- sum(!is.na(black_women_unarmed$race))
total_rows_female_unarmed <- sum(!is.na(female_unarmed$race))

#find percent change
percent(total_rows_black_women_unarmed / total_rows_female_unarmed)

#Black women accounted for 26.92% of unarmed deaths. This isn't the exact 28% figure, but we're unsure if this is a result of our syntax or just a database update since the time of publication.
```

**Q4: About 31 percent, or 77, of the 247 women fatally shot by police since 2015 had mental health issues, compared to 22 percent of the 5,362 men killed.**
<<<<<<< HEAD

first part of graf: about 31 percent, or 77, of the 247 women fatally shot by police since 2015 had mental health issues
```{r}
mental_health_women <- women_shootings %>% 
  filter(was_mental_illness_related == "TRUE")

count(mental_health_women)

#83 women had mental health issues 

total_women_mental_health <- sum(!is.na(mental_health_women))
total_women_shot <- sum(!is.na(women_shootings))

percent(total_women_mental_health / total_women_shot)

#This is slightly more than what it was previously, seeing as there are now 83 women recorded with mental health issues and 248 women recorded fatalities. Now, it's at about 33.48%.

#Since this database has been continuously updated since the publication of this investigation in 2020, there are likely to have been about six more women who were discovered to have mental health issues. 
=======
```{r}
#make a dataframe just for men
men_shot <- filtered_shootings %>% 
  filter(gender == "male")

count(men_shot)
#this count will also be off, because from this time it appears The Post discovered that there was actually one more man who was murdered before this investigation was published

#make a dataframe for men with mental health issues
men_mental_health <- men_shot %>% 
  filter(was_mental_illness_related == "True")

count(men_mental_health)

#calculate the percent total
total_men_mental_health <- sum(!is.na(men_mental_health))
men_shot_total <- sum(!is.na(men_shot))

percent(total_men_mental_health / men_shot_total)

#this total is also a little off, seeing as there have been more men discovered to have mental health issues since this investigation has been published. Instead of 22%, we see 23.25%. 
>>>>>>> 56260f4967ceb87a3c11b655042bb9d96dbecde7
```

second part of graf: 22 percent of the 5,362 men killed had mental health issues
```{r}
#make a dataframe for men
men_shot <- filtered_shootings %>% 
  filter(gender == "male")

count(men_shot)
#this count will also be off, because from this time it appears The Post discovered that there was actually one more man who was murdered before this investigation was published

#make a dataframe for men with mental health issues
men_mental_health <- men_shot %>% 
  filter(was_mental_illness_related == "True")
