------------------
name: JOUR472 draft reverse engineering data notebook
author: Olivia Borgula
date: 04/04/2024
------------------
```{r}
#turn off scientific notation
options(scipen = 999)
```

This project will reverse engineer findings by The Washington Post's story, headlined "Nearly 250 women fatally shot by police since 2015." The link is here: https://www.washingtonpost.com/graphics/2020/investigations/police-shootings-women/

This story describes the disparity in people fatally shot by police in regards to sex and within sex, race. We will replicate the following findings described in paragraphs in The Post’s story: 

“Since The Washington Post began tracking fatal shootings by police in 2015, officers have fatally shot 247 women out of the more than 5,600 people killed overall.”

“Of the 247 women fatally shot, 48 were Black and seven of those were unarmed. At least 89 of the women were at their homes or residences where they sometimes stayed. And 12 of those women killed at home were shot by officers who were there to conduct a search or make an arrest.”

“Since 2015, Black women have accounted for less than 1 percent of the overall fatal shootings in cases where race was known. But within this small subset, Black women, who are 13 percent of the female population, account for 20 percent of the women shot and killed and 28 percent of the unarmed deaths.”

“About 31 percent, or 77, of the 247 women fatally shot by police since 2015 had mental health issues, compared to 22 percent of the 5,362 men killed.”

But first, let’s load our libraries and data. 

```{r}
#load packages
library(tidyverse)
library(janitor)
library(dplyr)
library(formattable)
```

Here are the data sets The Washington Post used that we'll be using here. 
```{r}
#load the data
agencies <- read.csv("fatal-police-shootings-agencies.csv") %>% 
  clean_names()
shootings <- read.csv("fatal-police-shootings-data.csv") %>% 
  clean_names()
```

Now to be up to date, we're going to filter out any shootings that happened after Sept. 4, 2020, when this investigation was published.
```{r}
#filter for specific times that happened before sept. 4, 2020, when the story was published
filtered_shootings <- shootings %>%
  filter(date <= as.Date("2020-09-04"))

count(filtered_shootings)
#shows next part of the question, with more than 5,600 deaths
```

We are also now going to create a specific data frame just for women to find this part of the story: "Since The Washington Post began tracking fatal shootings by police in 2015, officers have fatally shot 247 women out of the more than 5,600 people killed overall.”

```{r}
#filter for women
women_shootings <- filtered_shootings%>%
  filter(gender == "female")

#count total women shot
count(women_shootings)

#not getting 247, but this is because the investigation was published four years ago and values have routinely been added. With this, ther emust have been another female death discovered during that period of time that caused this slight discrepancy. This is 248 out of more than 5,600, as shown in the code chunk before this one.
```

Now, to find this next part that we will break into parts: “Of the 247 women fatally shot, 48 were Black and seven of those were unarmed. At least 89 of the women were at their homes or residences where they sometimes stayed. And 12 of those women killed at home were shot by officers who were there to conduct a search or make an arrest.”

First, let's find this: “Of the 247 women fatally shot, 48 were Black and seven of those were unarmed."
*48 of the 247 women shot were Black.*
```{r}
#filter by race
race_women_shootings <- women_shootings %>% 
  filter(race == "B")

count(race_women_shootings)

#print(race_women_shootings)
```

*7 Black women were unarmed.*
```{r}
#out of the black women, find how many were unarmed
black_women_unarmed <- race_women_shootings %>% 
  filter(armed_with == "unarmed")

count(black_women_unarmed)
```

The findings about women being at home are impossible to replicate, seeing as the dataframes do not have columns specifically about the location of the victim and their addresses are not released. Therefore, we cannot reverse engineer this part of this paragraph: "At least 89 of the women were at their homes or residences where they sometimes stayed. And 12 of those women killed at home were shot by officers who were there to conduct a search or make an arrest.”


Next, let's find the information in this paragraph:
"About 31 percent, or 77, of the 247 women fatally shot by police since 2015 had mental health issues, compared to 22 percent of the 5,362 men killed."
```{r}
mental_health_women <- women_shootings %>% 
  filter(was_mental_illness_related == "True")

count(mental_health_women)

#Since this database has been continuously updated since the publication of this investigation in 2020, there are likely to have been about six more women who were discovered to have mental health issues. 
```
*About 31 percent, or 77, of the 247 women fatally shot by police since 2015 had mental health issues*
```{r}
total_women_mental_health <- sum(!is.na(mental_health_women))
total_women_shot <- sum(!is.na(women_shootings))

percent(total_women_mental_health / total_women_shot)

#This is slightly more than what it was previously, seeing as there are now 83 women recorded with mental health issues and 248 women recorded fatalities. Now, it's at about 33.48%.
  
```


*...compared to 22 percent of the 5,362 men killed.*
```{r}
#have to make a dataframe for just men
men_shot <- filtered_shootings %>% 
  filter(gender == "male")

count(men_shot)
#this count will also be off, because from this time it appears The Post discovered that there was actually one more man who was murdered before this investigation was published
```

```{r}
#also, now we have to make a dataframe for men with mental health issues
men_mental_health <- men_shot %>% 
  filter(was_mental_illness_related == "True")

count(men_mental_health)
```
```{r}
#now, let's actually do the calculation
total_men_mental_health <- sum(!is.na(men_mental_health))
men_shot_total <- sum(!is.na(men_shot))

percent(total_men_mental_health / men_shot_total)

#this total is also a little off, seeing as there have been more men discovered to have mental health issues since this investigation has been published. Instead of 22%, we see 23.25%. 
```
Lastly, let's find the information in this paragraph. 
"Since 2015, Black women have accounted for less than 1 percent of the overall fatal shootings in cases where race was known. But within this small subset, Black women, who are 13 percent of the female population, account for 20 percent of the women shot and killed and 28 percent of the unarmed deaths."

*Black women have accounted for less than 1 percent of the overall fatal shootings in cases where race was known*
```{r}
total_rows_bw <- sum(!is.na(race_women_shootings$race))
total_rows_shootings <- sum(!is.na(filtered_shootings$race))

#calculate the percent 
percent(total_rows_bw / total_rows_shootings)

#Black women account for 0.86% of the overall fatal shootings in cases where race was known, or less than 1 percent. We don't know if this too far from 1%, but is the value we received. 
```

*Black women account for 20 percent of the women shot and killed*
```{r}
#find number of rows in dataframe for Black women and female shootings dataframe with rows removed that contain "N/A" for race (assuming "where race was known" also applies to these sentences)
total_rows_black_women <- sum(!is.na(race_women_shootings$race))
total_rows_female <- sum(!is.na(women_shootings$race))

#calculate the percent
percent(total_rows_black_women / total_rows_female)

#Black women account for 19.35% of the women shot, or close to 20 percent - this figure is slightly off likely due to updates in the dataframe in the four years since the article was published. We're unsure as to why the number is lower, but it's unclear if the number changed since the investigation was published.
```
*Black women account for 28 percent of the unarmed deaths*
```{r}
#filter for unarmed females
female_unarmed <- filtered_shootings %>%
  filter(armed_with == "unarmed", gender == "female")
```

```{r}
#Find number of rows excluding those with "N/A" for race (black_women_unarmed doesn't have any "N/A" values for race, but adding based on mindset that wapo reporters are working with regularly updated data)
total_rows_black_women_unarmed <- sum(!is.na(black_women_unarmed$race))
total_rows_female_unarmed <- sum(!is.na(female_unarmed$race))

#find percent change
percent(total_rows_black_women_unarmed / total_rows_female_unarmed)

#Black women accounted for 26.92% of unarmed deaths. This isn't the exact 28% figure, but we're unsure if this is a result of our syntax or just a database update since the time of publication.
```





