#lab_08
Sean Mussenden, Aug 16, 2022

## About this lab

To complete this lab, you need to:
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings
**Task**: Load janitor and the tidyverse
```{r}
library(janitor)
library(tidyverse)
library(formattable)
library(dplyr)
```

## Load data

**Task**: Load four tables of U.S. Census data.  Each table has one row per Maryland county with 2020 population information.  Each table has these same two columns:


```{r}
#upload data
maryland_pop <- read_csv("data/maryland_population.csv")%>%
  clean_names()
maryland_hispanic <- read_csv("data/maryland_hispanic.csv")%>%
  clean_names()
maryland_white <- read_csv("data/maryland_white.csv")%>%
  clean_names()
maryland_black <- read_csv("data/maryland_black.csv")%>%
  clean_names()
```

* geoid - a unique identifier for each county
* name - the county name

The third column in each table represents the total population in one of four demographic categories.

* maryland_population.csv - total_pop - total population 
* maryland_hispanic.csv - hispanic_pop - total Hispanic population
* maryland_white.csv - white_pop - total White alone population
* maryland_black.csv - black_pop - total Black alone population

Some of the information in the tables may need to be cleaned or validated in order to produce the correct answer.  

## Questions

**Q1**  What Maryland county has the largest Hispanic population (by percentage of total population)? 
**A1** 
```{r}
#replace unclean data name for Montgomery County, join maryland_hispanic and maryland_pop and add a new column with Hispanic population percentage 
hispanic_pop_clean <- maryland_hispanic %>%
  mutate(clean_hispanic_pop = case_when(
    name == "Montgomery County,Maryland" ~ "Montgomery County, Maryland",
    TRUE ~ as.character(name)
  )) %>%
  select(name, clean_hispanic_pop, hispanic_pop)
updated_hispanic <- hispanic_pop_clean %>%
  inner_join(maryland_pop, by="name") %>%
  select(clean_hispanic_pop, hispanic_pop, total_pop)
largest_hispanic <- updated_hispanic %>%
  mutate(hisp_pct = percent(hispanic_pop / total_pop)) %>%
  arrange(desc(hisp_pct))

#Montgomery County has the largest Hispanic population at 18.79% of its population
```
**Q2**  What is the total population of people who identify as white alone in Maryland? 
**A2**  
```{r}
#replace incorrect Somerset value and find some of white alone population in cleaned data 
white_md_clean <- maryland_white %>%
  mutate(clean_white_pop = case_when(name == "Somerset County, Maryland" & geoid == 24039 ~ 13510, TRUE ~ white_pop))


sum(white_md_clean$clean_white_pop)
#3,275,048 people identify as white alone in Maryland
```

**Q3** What is the average Black population in all Maryland counties? 
**A3** 
```{r}
#replace N/As with data
maryland_black_clean <- maryland_black %>%
  mutate(clean_black_pop = case_when(
    name == "Baltimore city, Maryland" & geoid == 24510 ~ 358453,
    name == "Baltimore County, Maryland" & geoid == 24005 ~ 272596,
    TRUE ~ black_pop
  )) %>%
  select(geoid, name, clean_black_pop)

mean(maryland_black_clean$clean_black_pop)

#Average Black population for all Maryland counties is 75,670 people

```
