#lab_08
Sean Mussenden, Aug 16, 2022

## About this lab

To complete this lab, you need to:
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings
**Task**: Load janitor and the tidyverse and sf

```{r}
#load libraries
library(tidyverse)
library(janitor)
library(sf)
library(formattable)

#turn off scientific notation
options(scipen=999)
```


## Load data

You have been given two spatial dataframes and one regular dataframe to load.  The dataframes are:

* baltimore_homicides.rds | spatial | location and details about each Baltimore homicide between 2007 and 2017, including information about the status (open, closed with arrest, closed without arrest) and the exact location. Details on the data: https://github.com/washingtonpost/data-homicides
* baltimore_tract_shapefiles.rds | spatial | polygons of each census tract in Baltimore
* baltimore_tract_demo.rds | regular | the total population and white population of each baltimore Census tract in 2017.

**Task** Load the data

```{r}
baltimore_homicides <- read_rds("data/baltimore_homicides.rds")
baltimore_tract_demo <- read_rds("data/baltimore_tract_demo.rds")
baltimore_tract_shapefiles <- read_rds("data/baltimore_tract_shapefiles.rds")
```

## Questions

**Q1** Are homicides that occur in majority white Census tracts in Baltimore solved at a higher rate than homicides in non-majority white Baltimore census tracts?
* Write up your findings in three paragraphs that could be dropped into a news story, with the answer to the question as the lede. 
* Include any other relevant details.
* Output a table with relevant results. 
* Output at least one map visualization pertinent to your finding. 

**A1** 

```{r}
#assign each tract as either over or under 50% white (majority white)
majority_white <- baltimore_tract_demo %>%
  mutate(pct_white = white_pop/total_pop) %>%
  select(geoid, name, pct_white) %>%
  mutate(neighborhood_type = case_when(
    pct_white > .50 ~ "majority_white",
    TRUE ~ "not_majority_white"
  ))
```

```{r}
#join majority_white data with Baltimore tract shapefiles and convert to spatial object 
majority_white <- majority_white %>%
  inner_join(baltimore_tract_shapefiles) %>%
  st_as_sf()
```

```{r}
#assign each shooting to a tract with a spatial join
neighborhood_shootings <- baltimore_homicides %>%
  st_join(majority_white) %>%
  as_tibble()
```

```{r}
#find percent of unsolved homicides in white and non-white neighborhoods
case_status_neighborhoods <- neighborhood_shootings %>%
  group_by(neighborhood_type, disposition) %>%
  count() %>%
  pivot_wider(names_from = disposition, values_from = n) %>%
  clean_names() %>%
  mutate(total_shootings = closed_by_arrest + closed_without_arrest + open_no_arrest) %>%
  mutate(unsolved = open_no_arrest + closed_without_arrest) %>%
  mutate(pct_unsolved = percent(unsolved / total_shootings))

#table:
view(case_status_neighborhoods)

#majority white percent unsolved: 46.44%
#non-majority white percent unsolved: 66.23%

#calculate how many times higher the unsolved homicide rate is in non-majority white neighborhoods compared to majority white neighborhoods
.6623 / .4644

#1.43 times higher 
```

```{r}
#visualize this data 
ggplot() +
  geom_sf(data=baltimore_tract_shapefiles) +
  geom_sf(data=baltimore_homicides, color="navy")
```

**Paragraphs**
```{r}
#Homicides in majority white neighborhoods were solved at a higher rate than ones that occurred in non-majority white areas between 2007 and 2017. 

#About 44.4% of cases were unsolved in majority white neighborhoods compared to 66.2% in non-majority white neighborhoods — meaning cases were unsolved at a rate of about 1.43 times higher in non-majority white neighborhoods, according to a data analysis. 

#Out of 1,825 total unsolved homicides in that time frame, 111 occurred in majority white neighborhoods compared to 1,714 in non-majority white neighborhoods. 
```

